---
title: "L2L Framework: Data Preparation"
subtitle: "LAPOP Mexico 2023 & Latinobarómetro Colombia 2023"
author: "Kevin Linares (JPSM, University of Maryland)"
date: today
format:
  pdf:
    documentclass: article
    geometry: margin=1in
execute:
  echo: true
  warning: true
  message: false
---

```{r setup}
pacman::p_load(sjlabelled, psych, knitr, 
               viridis, patchwork, lavaan, semTools, tidyverse)

set.seed(721)

results_dir <- "D:/repos/UMD_classes_code/TSE_II_SURV721/results"
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Overview

This script prepares two survey datasets for the Latent-to-Language (L2L) framework:

-   **LAPOP Mexico 2023** — AmericasBarometer, 18-item trust/system support battery
-   **Latinobarómetro Colombia 2023** — 9-item institutional trust battery

For each survey, we:

-   \(1\) load raw data

-   \(2\) inspect value labels to understand coding

-   \(3\) select and recode variables

-   \(4\) assess missingness and drop high-missing demographics

-   \(5\) apply listwise deletion

-   \(6\) validate the final dataset.

# LAPOP Mexico 2023

## Load and Inspect

```{r lapop-load}
lapop_raw <- read_spss(
  "D:/repos/UMD_classes_code/TSE_II_SURV721/lapop/MEX_2023_LAPOP_AmericasBarometer_v1.0_w.sav") |>
  tibble()

```

### Trust items "Trust in"

Coded as 1 = Not at all, . . ., 7 = a lot.

```{r lapop-trust-inspect}
trust_items <- c("b10a", "b11", "b12", "b13", "b15",
                 "b18", "b19", "b21", "b21a", "b31",
                 "b32", "b37", "b47a")


lapop_trust_names <- c(
  justice_system = "b10a",
  electoral_tribunal = "b11",
  armed_forces = "b12",
  legislature = "b13",
  public_ministry = "b15",
  police = "b18",
  auditor = "b19",
  political_parties = "b21",
  president = "b21a",
  supreme_court = "b31",
  municipality = "b32",
  media = "b37",
  elections = "b47a"
)


lapop_raw |>
  select(all_of(trust_items)) |>
  rename(!!!lapop_trust_names) |> 
  drop_na() |> 
  pivot_longer(everything(), names_to = "item", values_to = "value") |>
  count(item, value) |>
  ggplot(aes(x = item, y = n, fill = factor(value))) +
  geom_col(position = "fill") +
  scale_fill_viridis_d() +
  labs(x = NULL, y = "Proportion", fill = "Response") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

Trust items

### Outcome Variables

```{r lapop-outcome-inspect}

lapop_raw |>
  select(pn4, m1) |>
  mutate_all(sjlabelled::as_character) |>
  pivot_longer(everything(), 
               names_to = "Outcome", values_to = "Response") |> 
  drop_na() |> 
  group_by(Outcome, Response) |> 
  count() |>
  group_by(Outcome) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = Response, y = pct, fill = Response)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.5) +
  facet_wrap(~Outcome, scales = "free_x") +
  labs(y = "Percentage", x = NULL) +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

### M1 is presidential rating. The following are the labels translated from Spanish.

-   1 = Very good

-   2 = Good

-   3 = Neighter good nor bad

-   4 = Bad

-   5 = Very Bad

### pn4 is satisfaction with democracy

-   1 = Very satisfied

-   2 = satisfied

-   3 = dissatisfied

-   4 = very dissatisfied

## Recode data

```{r lapop-trust-recode}
# demographics 
lapop_demo_source <- c("q1tc_r", "q2", "edre", "ur", "ocup4a")

lapop <- lapop_raw |>
  select(all_of(c(trust_items, lapop_demo_source, "pn4", "m1"))) |>
  rename(!!!lapop_trust_names) |> 
  mutate(across(all_of(setdiff(lapop_demo_source, "q2")), as_character)) |>   
  # rename variables & recode values to english
  mutate(
    # bin age
     age_cat = cut(q2,
                  breaks = c(17, 29, 44, 59, Inf),
                  labels = c("18-29", "30-44", "45-59", "60+"),
                  right = TRUE),
     
    # recode gender to factor with female as reference group
    sex = ifelse(q1tc_r == "Mujer/femenino", "Female", "Male"),
    male = factor(sex),
    
    # recode education
    Edu = case_when(
      edre == "Ninguna" ~ "none",
      edre == "Primaria incompleta" ~ "primary",
      edre == "Primaria completa" ~ "primary",
      edre == 
        "Secundaria o Educación Media Superior/Bachillerato/Preparatoria/Profesional Técnico incompleta" ~ "secondary",
      edre == 
        "Secundaria o Educación Media Superior/Bachillerato/Preparatoria/Profesional Técnico completa" ~ "secondary",
      edre == 
        "Universitaria, superior no universitaria o técnico universitario completa" ~ "higher_ed",
      edre == 
        "Universitaria, superior no universitaria o técnico universitario incompleta" ~ "higher_ed",
      TRUE ~ NA_character_
    ),
    Edu = factor(Edu, 
                 levels = c("none", "primary", "secondary", "higher_ed")),
    
    # urban factor
    Urban = factor(ifelse(ur=="Urbano", "urban", "rural")),
    
    # recode occupation status
    Employment = case_when(
      ocup4a == "Trabajando?" ~ "Employed",
      ocup4a == "No está trabajando en este momento pero tiene trabajo?" 
      ~ "Employed",
      ocup4a == "Se dedica a los quehaceres de su hogar?" ~ "house_wife",
      ocup4a == "Es estudiante?" ~ "Student",
      ocup4a == 
        "Está jubilado, pensionado o incapacitado permanentemente para trabajar?" ~ "Retired",
      ocup4a == "Está buscando trabajo activamente?" ~ "Unemployed",
      ocup4a == "No trabaja y no está buscando trabajo?" ~ "Unemployed",
      TRUE ~ NA_character_),
    Employment = factor(Employment),
    
    # recode outcomes
    satis_demo_lapop = pn4,
    prez_rating = m1
    ) |> 
  # listwise deletion
  drop_na() |> 
  # drop variables we no longer need
  select(!all_of(lapop_demo_source), -sex, -pn4, -m1)

lapop |> glimpse()


map(1:ncol(lapop), function(C) { 
  col_name <- names(lapop)[C]
  lapop |> select(all_of(C)) |> rename(Var=1) |> count(Var) |> mutate(column = col_name, .before = 1) |> 
    kable()
})

```

# We are left with n=1,430

### 

## 

## 

\newpage

# Latinobarometro Colombia 2023

## Load data

```{r}
lb_raw <- read_rds(
  "D:/repos/UMD_classes_code/TSE_II_SURV721/latinobarrometer/Latinobarometro_2023_Esp_Rds_v1_0.rds") |>
  tibble() |>
  # keep only Colombia
  filter(idenpa == 170)
```

### Trust Battery (9 items) and responses translated to English.

-   -5 through -1 = NA

-   1 = A lot of trust

-   2 = some trust

-   3 = little trust

-   4 = no trust

```{r}
trust_vars_raw <- c("P13STGBS_A", "P13STGBS_B", "P13ST_C",
                     "P13ST_D", "P13ST_E", "P13ST_F",
                     "P13ST_G", "P13ST_H", "P13ST_I")

lb_trust_names <- c(Armed_forces = "P13STGBS_A",
                    Police = "P13STGBS_B", 
                    Church = "P13ST_C",
                    Congress = "P13ST_D", 
                    Natl_Govt = "P13ST_E", 
                    Judiciary = "P13ST_F",
                    Parties = "P13ST_G",
                    Electoral_Inst = "P13ST_H", 
                    President = "P13ST_I")


lb_raw |>
  select(all_of(trust_vars_raw)) |>
  rename(!!!lb_trust_names) |>
  mutate(across(everything(), ~ifelse(. %in% -5:-1, NA, .))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  drop_na(value) |>
  count(variable, value) |>
  group_by(variable) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = variable, y = prop, fill = factor(value))) +
  geom_col() +
  scale_fill_viridis_d() +
  labs(x = "Variable", y = "Proportion", fill = "Response") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Outcome Variables

### P11STGBS_A Satisfaction with Democracy codings translated into English.

-   -5 through -1 = NA

-   1 = Very Satisfied

-   2 = Quite Satisfied

-   3 = Not very satisfied

-   4 = not at all satisfied

### P12ST country governed by a few powerful groups, or governed for the good at all codings translated into English

-   -5 through -1 = NA

-   1 = powerful groups in their interest

-   2= for the good of all

```{r}

lb_raw |> 
  select(P11STGBS_A, P12ST) |> 
  mutate(across(everything(), ~if_else(. < 0, NA_real_, .))) |> 
  mutate(across(everything(), factor)) |> 
  pivot_longer(everything(), names_to = "variable", values_to = "value") |> 
  ggplot(aes(x = value)) +
  geom_bar(aes(fill = variable), show.legend = FALSE) +
  facet_wrap(~variable, scales = "free") +
  scale_fill_viridis_d() +
  theme_minimal()


```

### 

### 

## Recode data

```{r}

# demographics 
lb_demo_source <- c("sexo", "edad",
                     "REEEDUC_1", "tamciud", "S18_A")

lb <- lb_raw |> 
  select(all_of(trust_vars_raw), all_of(lb_demo_source), P11STGBS_A, P12ST) |> 
  rename(!!!lb_trust_names) |>
  mutate(across(all_of(setdiff(lb_demo_source, "edad")), as_character)) |> 
  
  # remove missing data
  mutate(across(everything(), ~ifelse(. %in% -5:-1, NA, .)),
         
         # recode demographics
          male = factor(
            ifelse(sexo == "Hombre", "Male", "Female")),
         
         # bin age
         age_cat = cut(edad,
                  breaks = c(17, 29, 44, 59, Inf),
                  labels = c("18-29", "30-44", "45-59", "60+"),
                  right = TRUE),
         
         # recode education
    Edu = case_when(
      REEEDUC_1 == "Analfabeto" ~ "none",
      REEEDUC_1 == "Basica incompleta" ~ "primary",
      REEEDUC_1 == "Basica completa" ~ "primary",
      REEEDUC_1 =="Secundaria, media, tecnica completa" ~ "secondary",
      REEEDUC_1 == "Secundaria, media, tecnica incompleta" ~ "secondary",
      REEEDUC_1 == "Superior completa" ~ "higher_ed",
      REEEDUC_1 == "Superior Incompleta" ~ "higher_ed",
      TRUE ~ NA_character_
    ),
    Edu = factor(Edu, 
                 levels = c("none", "primary", "secondary", "higher_ed")),
         
    # urban factor
    Urban = case_when(
      tamciud == "(Capital)" ~ "urban",
      tamciud == "100.001 y mas" ~ "urban",
      tamciud == "50.001 - 100.000" ~ "urban",
      tamciud == "40.001 - 50.000" ~ "urban",
      tamciud == "20.001 - 40.000" ~ "rural",
      tamciud == "10.001 - 20.000" ~ "rural",
      tamciud == "5.001 - 10.000" ~ "rural",
      tamciud == "Menos de 5.000" ~ "rural",
      TRUE ~ NA_character_
    ),
    
    Urban = factor(Urban, levels = c("rural", "urban")),
    
    # recode occupation status
    Employment = case_when(
      S18_A == "Independiente/cuenta propia" ~ "Employed",
      S18_A == "Asalariado en emp. privada" ~ "Employed",
      S18_A == "Asalariado en emp. publica" ~ "Employed",
      S18_A == 
        "No trabaja/ responsable de las compras y el cuidado de la casa" ~
        "house_wife",
      S18_A == "Estudiante" ~ "Student",
      S18_A ==  "Retirado/pensionado" ~ "Retired",
      S18_A == "Temporalmente no trabaja" ~ "Unemployed",
      TRUE ~ NA_character_),
    Employment = factor(Employment),
    
    # recode outcomes
    satis_demo_lb = P11STGBS_A,
    governed_for = P12ST
    ) |> 

  # listwise deletion
  drop_na() |> 
  
  # drop variables we no longer need
  select(!all_of(lb_demo_source), -P11STGBS_A, -P12ST)

lb |> glimpse()


map(1:ncol(lb), function(C) { 
  col_name <- names(lb)[C]
  lb |> select(all_of(C)) |> rename(Var=1) |> count(Var) |> mutate(column = col_name, .before = 1) |> 
    kable()
})

```

# We are left with n=1,073

## 

\newpage

## We now move to exploratory factor analysis for each survey trust items.

### Lapop EFA with one-three factors

```{r}
# ══════════════════════════════════════════════════════════════════════════════
# Exploratory Factor Analysis — Reusable Function
# ══════════════════════════════════════════════════════════════════════════════

run_efa <- function(items_df, max_factors = 3, seed = 721) {

  # Coerce to numeric, drop incomplete cases
  items_df <- items_df |> drop_na() |> mutate(across(everything(), as.numeric))
  cat(sprintf("EFA: %d cases, %d items\n\n", nrow(items_df), ncol(items_df)))

  # Polychoric correlations — appropriate for ordinal items
  R <- polychoric(items_df)$rho

  # Parallel analysis — compare observed eigenvalues to simulated null
  set.seed(seed)
  pa <- fa.parallel(R, n.obs = nrow(items_df), fa = "fa", cor = "cor",
                    plot = TRUE, main = "Parallel Analysis")

  # Fit 1 through max_factors — oblimin rotation for correlated factors
  models <- map(1:max_factors, \(nf) {
    fa(R, nfactors = nf, n.obs = nrow(items_df),
       rotate = if (nf == 1) "none" else "oblimin",
       fm = "wls", cor = "cor")
  }) |> set_names(paste0("f", 1:max_factors))

  # Comparison table — TLI, RMSEA, BIC for model selection
  comparison <- imap_dfr(models, \(mod, name) {
    tibble(model = name, factors = mod$factors,
           TLI = mod$TLI, RMSEA = mod$RMSEA[1],
           BIC = mod$BIC,
           prop_var = sum(mod$Vaccounted["Proportion Var", ]))
  })
  cat("\n"); print(comparison, width = Inf)

  # Print sorted loadings (cutoff = 0.3) for each solution
  walk(models, \(mod) {
    cat(sprintf("\n-- %d-factor loadings --\n", mod$factors))
    print(mod$loadings, cutoff = 0.3, sort = TRUE)
  })

  # Return key objects for downstream CFA specification
  invisible(list(polychoric = R, parallel = pa, models = models,
                 comparison = comparison, n_suggested = pa$nfact))
}
```

```{r}
lapop_trust <- run_efa(lapop |> select(justice_system:elections), max_factors = 3)
```

-   We find that a one factor solution is feasible but has a poor fit indicated by TLI and RMSEA. It seems that trust in the president might be a problematic item.

### We now conduct a CFA one and two factor model on the lapop trust items

```{r}
# ══════════════════════════════════════════════════════════════════════════════
# CFA — 1-Factor and 2-Factor Models (LAPOP Mexico Trust Items)
# ══════════════════════════════════════════════════════════════════════════════

set.seed(721)

diagnose_cfa <- function(fit, label) {

  cat(sprintf("\n%s\n  %s\n%s\n", strrep("=", 60), label, strrep("=", 60)))

  fm <- fitmeasures(fit, c("chisq.scaled", "df.scaled", "pvalue.scaled",
                            "cfi.scaled", "tli.scaled",
                            "rmsea.scaled", "rmsea.ci.lower.scaled",
                            "rmsea.ci.upper.scaled", "srmr"))
  cat(sprintf("\nchi2 = %.1f (df = %.0f, p = %.4f)\n", fm[1], fm[2], fm[3]))
  cat(sprintf("CFI = %.3f | TLI = %.3f | RMSEA = %.3f [%.3f, %.3f] | SRMR = %.3f\n",
              fm[4], fm[5], fm[6], fm[7], fm[8], fm[9]))

  std <- standardizedSolution(fit) |>
    filter(op == "=~") |>
    select(factor = lhs, item = rhs, loading = est.std, p = pvalue) |>
    filter(!is.na(loading)) |>
    arrange(factor, desc(abs(loading)))
  cat("\nLoadings:\n"); print(as.data.frame(std))

  fcor <- standardizedSolution(fit) |>
    filter(op == "~~", lhs != rhs, lhs %in% std$factor, rhs %in% std$factor)
  if (nrow(fcor) > 0)
    cat(sprintf("\nFactor correlation: %.3f\n", fcor$est.std[1]))

  omega <- tryCatch(compRelSEM(fit), error = function(e) NULL)
  ave   <- tryCatch(AVE(fit), error = function(e) NULL)
  if (!is.null(omega)) cat(sprintf("omega: %s\n",
    paste(names(omega), round(omega, 3), sep = " = ", collapse = ", ")))
  if (!is.null(ave)) cat(sprintf("AVE:   %s\n",
    paste(names(ave), round(ave, 3), sep = " = ", collapse = ", ")))

  mi <- modificationIndices(fit, sort. = TRUE, op = "~~", free.remove = FALSE) |>
    filter(mi > 10) |>
    slice_head(n = 5)
  if (nrow(mi) > 0) {
    cat("\nTop modification indices (> 10):\n")
    print(as.data.frame(mi |> select(lhs, rhs, mi)))
  }

  invisible(fit)
}

```

### We find that the president trust item is problematic and test a 1 factor cfa without. 

```{r}

trust_items_lapop <- c("justice_system", "electoral_tribunal", "armed_forces",
                     "legislature", "public_ministry", "police", "auditor",
                     "political_parties", "supreme_court", "municipality",
                     "media", "elections")

trust_lapop_df <- lapop |> select(all_of(trust_items_lapop)) 

# 1-factor, 12 items, freed residual covariances
mod_1f_trim <- '
  Trust =~ justice_system + electoral_tribunal + armed_forces +
           legislature + public_ministry + police + auditor +
           political_parties + supreme_court + municipality +
           media + elections
  armed_forces ~~ legislature
  public_ministry ~~ auditor
'

fit_1f_trim <- cfa(mod_1f_trim, data = trust_lapop_df, 
                    ordered = setdiff(trust_items_lapop, "president"),
                    estimator = "WLSMV")

diagnose_cfa(fit_1f_trim, "1-FACTOR (12 items, 2 freed covariances)")
```

#### **12-item unidimensional Institutional Trust factor**, with two freed residual covariances (armed_forces \~\~ legislature, public_ministry \~\~ auditor), president dropped. This is what the data support — the EFA said one dominant factor, the CFA confirms it once the misfit item is removed and substantively justified correlated errors are freed. This will be our production script.

We move to estimating the correlation between the latent factor and our outcome variables.

```{r}

# Factor scores
fs <- lavPredict(fit_1f_trim, method = "regression") |> as_tibble()

lapop_fs <- lapop |>
  filter(if_all(all_of(trust_items_lapop), ~ !is.na(.x))) |>
  bind_cols(fs)

# Correlations
cor(lapop_fs$Trust,
    lapop_fs |> select(satis_demo_lapop, prez_rating) |> mutate(across(everything(), as.numeric)),
    use = "complete.obs")
```

\newpage

## We now explore the factor structure of the LB data.

```{r}
lb_trust <- run_efa(lb |> select(Armed_forces:President), max_factors = 3)
```

The Latinobarómetro trust items are clearly not unidimensional. The 1-factor model is unacceptable (TLI = .636, RMSEA = .215), meaning these 9 items cannot be treated as measuring a single construct. The 3-factor solution provides the cleanest simple structure: Factor 1 is political trust (Natl_Govt .85, President .80, Parties .71, Congress .68, Electoral_Inst .66, Judiciary .65), Factor 2 is security institutions (Armed_forces .87, Police .78), and Factor 3 is Church alone (.85). Factor correlations are moderate and distinct (.45, .34, .20), confirming these are separable dimensions. The 2-factor solution is messy — Congress, Judiciary, Parties, and Electoral_Inst all cross-load heavily (complexity \~1.9), meaning the 2-factor structure forces ambiguous assignments. Church is the problem item: it loads .29 on the general factor, forms its own single-item factor in the 3-factor solution, and has near-zero correlations with President (.03) and Natl_Govt (.08). It is substantively distinct from institutional trust and should be dropped. That leaves a 2-factor CFA: Political Trust (6 items) and Security Trust (Armed_forces, Police), with Natl_Govt \~\~ President (r = .81 in the polychoric matrix) as a likely candidate for a freed residual covariance.

```{r}
# ══════════════════════════════════════════════════════════════════════════════
# CFA — Latinobarómetro Colombia (8 items, Church dropped)
# ══════════════════════════════════════════════════════════════════════════════

set.seed(721)

trust_items_lb <- c("Armed_forces", "Police", "Congress", "Natl_Govt",
                     "Judiciary", "Parties", "Electoral_Inst", "President")

trust_lb_df <- lb |> select(all_of(trust_items_lb)) 

# ── 1-factor (baseline) ─────────────────────────────────────────────────────
mod_lb_1f <- '
  Trust =~ Armed_forces + Police + Congress + Natl_Govt +
           Judiciary + Parties + Electoral_Inst + President
'

# ── 2-factor (EFA-informed) ─────────────────────────────────────────────────
mod_lb_2f <- '
  Political =~ Congress + Natl_Govt + Judiciary + Parties +
               Electoral_Inst + President
  Security  =~ Armed_forces + Police
'

# ── 2-factor with freed covariance ──────────────────────────────────────────
mod_lb_2f_free <- '
  Political =~ Congress + Natl_Govt + Judiciary + Parties +
               Electoral_Inst + President
  Security  =~ Armed_forces + Police
  Natl_Govt ~~ President
'

# ── Fit all three ────────────────────────────────────────────────────────────
fit_lb_1f     <- cfa(mod_lb_1f, data = trust_lb_df,
                      ordered = trust_items_lb, estimator = "WLSMV")
fit_lb_2f     <- cfa(mod_lb_2f, data = trust_lb_df,
                      ordered = trust_items_lb, estimator = "WLSMV")
fit_lb_2f_free <- cfa(mod_lb_2f_free, data = trust_lb_df,
                       ordered = trust_items_lb, estimator = "WLSMV")


# ── Diagnose ─────────────────────────────────────────────────────────────────
diagnose_cfa(fit_lb_1f, "LB 1-Factor (8 items)")
diagnose_cfa(fit_lb_2f, "LB 2-Factor: Political + Security")
diagnose_cfa(fit_lb_2f_free, "LB 2-Factor + freed Natl_Govt ~~ President")

# ── LRT comparison ───────────────────────────────────────────────────────────
lavTestLRT(fit_lb_1f, fit_lb_2f)
```

The 2-factor with freed Natl_Govt \~\~ President is the Colombia measurement model. CFI = .988, TLI = .981, RMSEA = .071, SRMR = .033 — all criteria pass. The factor correlation at .550 is moderate, confirming Political and Security are genuinely distinct dimensions (unlike the LAPOP 2-factor at .910). The LRT confirms the 2-factor is significantly better than 1-factor.

```{r}
fs_lb <- lavPredict(fit_lb_2f_free, method = "regression") |> as_tibble()

lb_fs <- lb |>
  filter(if_all(all_of(trust_items_lb), ~ !is.na(.x))) |>
  bind_cols(fs_lb)

cor(lb_fs |> select(Political, Security),
    lb_fs |> select(satis_demo_lb, governed_for) |> mutate(across(everything(), as.numeric)),
    use = "complete.obs")
```

**Political Trust → satis_demo_lb (r = .37):** Higher factor score (less trust) → higher satis_demo values (less satisfied). Equivalently, people who trust political institutions more are more satisfied with democracy. Intuitive direction, moderate-strong relationship.

**Political Trust → governed_for (r = -.29):** Higher factor score (less trust) → lower governed_for (1 = powerful groups). People with less political trust think the country is governed by powerful groups. Intuitive direction, moderate relationship.

**Security → satis_demo_lb (r = .21):** Same direction as Political but weaker. Security institution trust matters less for democratic satisfaction.

**Security → governed_for (r = -.07):** Essentially zero. Whether you trust the armed forces and police has almost no bearing on whether you think the country is governed for everyone's benefit. This makes substantive sense — governance perceptions are about political power, not security institutions.

We drop the security factor and find cleaner results.

\newpage

|   | LAPOP Mexico | Latinobarómetro Colombia |
|----|----|----|
| **Items** | 12 (of 13, president dropped) | 6 (of 9, Church/Armed_forces/Police dropped) |
| **Factors** | 1 (Trust) | 1 (Political) |
| **Freed covariances** | armed_forces \~\~ legislature, public_ministry \~\~ auditor | Natl_Govt \~\~ President |
| **CFI / TLI** | .964 / .955 | .992 / .985 |
| **RMSEA** | .056 | .077 |
| **omega** | .922 | .793 |
| **AVE** | .521 | .551 |
| **→ satis_demo** | r = -.33 | r = .37 |
| **→ prez_rating / governed_for** | r = -.26 | r = -.30 |

```{r}
# ══════════════════════════════════════════════════════════════════════════════
# Final CFA Models — Both Surveys
# ══════════════════════════════════════════════════════════════════════════════

# ── LAPOP Mexico: 12-item unidimensional Trust ───────────────────────────────
# President dropped (weak loading, single-item EFA factor, misfit source)
# Two freed residual covariances (substantively justified):
#   armed_forces ~~ legislature    (state power institutions)
#   public_ministry ~~ auditor     (oversight/accountability bodies)

lapop_cfa <- '
  Trust =~ justice_system + electoral_tribunal + armed_forces +
           legislature + public_ministry + police + auditor +
           political_parties + supreme_court + municipality +
           media + elections
  armed_forces ~~ legislature
  public_ministry ~~ auditor
'

# ── Latinobarómetro Colombia: 6-item unidimensional Political Trust ──────────
# Church dropped (h2 = .085, near-zero correlations with political items)
# Armed_forces and Police dropped (Security factor uncorrelated with outcomes)
# One freed residual covariance (substantively justified):
#   Natl_Govt ~~ President  (polychoric r = .81, conceptual overlap)

lb_cfa <- '
  Political =~ Congress + Natl_Govt + Judiciary + Parties +
               Electoral_Inst + President
  Natl_Govt ~~ President
'
```

## To save data into a .rds we will combine both surveys into a list wtih two elements.

```{r}

survey_list <- list(
  # drop president variable
  lapop |> select(-president), 
  # drop armed forces, police, and church
  lb |> select(-Armed_forces:-Church)) |> 
  set_names(c("lapop", "lb"))

write_rds(survey_list,
          file = str_c(
            results_dir, "/survey_data.rds")
          )
```
