---
title: "L2L Framework: 5-Fold Cross-Validation, CFA, IRT, and Demographic Profiling"
subtitle: "Script 02 — Psychometric Infrastructure for Item Imputation"
author: "Kevin Linares"
date: today
format:
  pdf:
    colorlinks: true
    fig-pos: "H"
    geometry: margin=1in
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  echo: true
  warning: false
  message: false
  cache: false
editor:
  markdown:
    wrap: 72
---

# Overview

This script builds the complete psychometric infrastructure for the
L2L item imputation experiment. All downstream scripts load the
outputs produced here to ensure every method is evaluated on identical
held-out cases with identical psychometric profiles.

Four analysis components are fitted per fold:

1. **CFA:** Unidimensional trust factor (ML estimation). Provides
   factor scores, loadings, fit indices, and composite reliability.
2. **GRM/IRT:** Unidimensional Graded Response Model with SEs.
   Provides discrimination parameters, Q3 diagnostics, theta
   estimates.
3. **Demographic Profiling:** For each demographic variable, computes
   (a) eta-squared with theta to screen which demographics carry
   meaningful signal, (b) per-item group means for direction and
   magnitude, (c) model-based conditional probabilities
   $P(X_j \geq k \mid \bar{\theta}_{\text{group}})$ for prompt
   context, and (d) conditional response distributions
   $P(X_j = k \mid \text{group})$ for the strongest associations.
4. **Masking Table:** Random assignment of 1--4 masked items per
   test respondent, shared across all downstream scripts.

## Changes from Previous Version

- **Dropped:** Measurement invariance (MI) and DIF (lordif). The
  diagnostic analysis showed DIF R² values below 0.03 across all
  items, meaning DIF flags added prompt tokens without actionable
  signal. The lordif dependency is removed.
- **Dropped:** Theta regression. Replaced by demographic profiling
  which produces the same information (demographic × latent
  relationship) in a format directly usable by LLM prompts:
  model-based probabilities rather than regression coefficients.
- **Added:** Employment as the fifth demographic variable, resolving
  the mismatch between Scripts 03 and 04.
- **Added:** Demographic screening via eta-squared threshold,
  computed per survey independently.
- **Added:** Model-based probability context per demographic group,
  transforming IRT parameters into $P(X_j \geq k \mid \theta)$
  that LLMs can interpret without psychometric expertise.

## Key Design Decisions

- **5-fold CV** (not 10) for larger training folds (~80% of N),
  required for stable CFA estimation.
- **IRT reverse-coding for LB** so higher theta = higher trust.
- **Demographic screening threshold:** eta² ≥ 0.005 with theta.
  Demographics below this threshold are excluded from prompts for
  that survey to avoid adding noise.
- **Same five demographics tested** for both surveys; survivors
  determined independently per survey.

# Setup

```{r}
#| label: setup

library(rsample)
library(lavaan)
library(semTools)
library(mirt)
library(ggridges)
library(viridis)
library(tidyverse)

set.seed(721)

results_dir <- "D:/repos/UMD_classes_code/TSE_II_SURV721/results"
survey_list <- readRDS(file.path(results_dir, "survey_data.rds"))
```

# Survey Configuration

```{r}
#| label: config

config <- list(
  lapop = list(
    data            = survey_list$lapop,
    demo_vars       = c("male", "age_cat", "Edu", "Urban",
                        "Employment"),
    trust_items     = c(
      "justice_system", "electoral_tribunal", "armed_forces",
      "legislature", "public_ministry", "police", "auditor",
      "political_parties", "supreme_court", "municipality",
      "media", "elections"
    ),
    cfa_syntax = '
      Trust =~ justice_system + electoral_tribunal + armed_forces +
               legislature + public_ministry + police + auditor +
               political_parties + supreme_court + municipality +
               media + elections
      armed_forces ~~ legislature
      public_ministry ~~ auditor
    ',
    factor_names    = "Trust",
    trust_ascending = TRUE
  ),
  lb = list(
    data            = survey_list$lb,
    demo_vars       = c("male", "age_cat", "Edu", "Urban",
                        "Employment"),
    trust_items     = c(
      "Congress", "Natl_Govt", "Judiciary", "Parties",
      "Electoral_Inst", "President"
    ),
    cfa_syntax = '
      Political =~ Congress + Natl_Govt + Judiciary + Parties +
                   Electoral_Inst + President
      Natl_Govt ~~ President
    ',
    factor_names    = "Political",
    trust_ascending = FALSE
  )
)

# Demographic screening threshold
ETA_SQ_THRESHOLD <- 0.005

# Q3 threshold for local dependence
Q3_THRESHOLD <- 0.10

# Number of folds
N_FOLDS <- 5L
```

# Helper Functions

```{r}
#| label: fn-helpers

# Label trust level relative to training distribution
label_trust <- function(fs_test, fs_train) {
  m <- mean(fs_train)
  s <- sd(fs_train)
  case_when(
    fs_test < (m - 1.0 * s) ~ "VERY LOW",
    fs_test < (m - 0.5 * s) ~ "LOW",
    fs_test < (m + 0.5 * s) ~ "MODERATE",
    fs_test < (m + 1.0 * s) ~ "HIGH",
    TRUE                     ~ "VERY HIGH"
  )
}

# Verbal labels for observed responses
lapop_trust_labels <- c(
  "1" = "No trust at all", "2" = "Very little trust",
  "3" = "Little trust",    "4" = "Some trust",
  "5" = "Moderate trust",  "6" = "High trust",
  "7" = "A lot of trust"
)

lb_trust_labels <- c(
  "1" = "A lot of trust", "2" = "Some trust",
  "3" = "Little trust",   "4" = "No trust"
)

# Eta-squared from one-way ANOVA
compute_eta_sq <- function(y, x) {
  valid <- !is.na(y) & !is.na(x)
  y <- y[valid]; x <- x[valid]
  if (length(unique(x)) < 2 || length(y) < 10) {
    return(tibble(eta_sq = NA_real_, F_stat = NA_real_,
                  p_value = NA_real_, n = length(y)))
  }
  fit <- aov(y ~ x)
  ss  <- summary(fit)[[1]]
  tibble(
    eta_sq  = round(ss[["Sum Sq"]][1] / sum(ss[["Sum Sq"]]), 4),
    F_stat  = round(ss[["F value"]][1], 2),
    p_value = round(ss[["Pr(>F)"]][1], 4),
    n       = length(y)
  )
}
```

# Build 5-Fold CV Splits

```{r}
#| label: build-folds

build_survey_folds <- function(cfg, survey_name) {
  full_data <- cfg$data |>
    mutate(row_id = row_number())

  cv_obj <- vfold_cv(full_data, v = N_FOLDS)

  map_dfr(seq_len(nrow(cv_obj)), function(fold_i) {
    sp <- cv_obj$splits[[fold_i]]
    bind_rows(
      training(sp) |> mutate(split = "train"),
      testing(sp)  |> mutate(split = "test")
    ) |>
      mutate(survey = survey_name, fold_id = fold_i) |>
      select(survey, fold_id, split, row_id, everything())
  })
}

folds_list <- list(
  lapop = build_survey_folds(config$lapop, "lapop"),
  lb    = build_survey_folds(config$lb,    "lb")
)

walk(names(folds_list), function(s) {
  cat(s, ":", nrow(folds_list[[s]]), "rows x",
      ncol(folds_list[[s]]), "columns\n")
  folds_list[[s]] |>
    count(fold_id, split) |>
    pivot_wider(names_from = split, values_from = n) |>
    print()
  cat("\n")
})
```

# Component 1: Per-Fold CFA

```{r}
#| label: fn-cfa

fit_fold_cfa <- function(train_df, test_df, cfg, survey, fold_id) {

  factor_name  <- cfg$factor_names
  label_lookup <- if (factor_name == "Trust") {
    lapop_trust_labels
  } else {
    lb_trust_labels
  }
  trust_items <- cfg$trust_items

  trust_train <- train_df |>
    select(all_of(trust_items)) |>
    mutate(across(everything(), as.numeric))
  trust_test <- test_df |>
    select(all_of(trust_items)) |>
    mutate(across(everything(), as.numeric))

  item_min <- min(unlist(trust_train), na.rm = TRUE)
  item_max <- max(unlist(trust_train), na.rm = TRUE)

  cfa_fit <- cfa(cfg$cfa_syntax, data = trust_train,
                  estimator = "ML", meanstructure = TRUE)

  params_df <- parameterEstimates(cfa_fit)
  load_df   <- params_df |> filter(op == "=~")
  intcpt_df <- params_df |> filter(op == "~1",
                                    lhs %in% trust_items)

  fs_train <- as.numeric(
    lavPredict(cfa_fit, newdata = trust_train,
               method = "regression"))
  fs_test <- as.numeric(
    lavPredict(cfa_fit, newdata = trust_test,
               method = "regression"))

  ecdf_fn      <- ecdf(fs_train)
  pctile_train <- round(ecdf_fn(fs_train) * 100)
  pctile_test  <- round(ecdf_fn(fs_test)  * 100)
  label_train  <- label_trust(fs_train, fs_train)
  label_test   <- label_trust(fs_test,  fs_train)

  # Expected item responses from CFA
  compute_expected <- function(fs_vec) {
    exp_cols <- map(trust_items, function(item) {
      ld  <- load_df$est[load_df$rhs == item]
      int <- intcpt_df$est[intcpt_df$lhs == item]
      stopifnot(length(ld) == 1, length(int) == 1)
      raw_val   <- int + ld * fs_vec
      truncated <- pmax(item_min, pmin(item_max, raw_val))
      rounded   <- as.integer(round(truncated))
      verbal    <- unname(label_lookup[as.character(rounded)])
      tibble(!!paste0("exp_", item) := verbal)
    })
    list_cbind(exp_cols)
  }

  exp_train <- compute_expected(fs_train)
  exp_test  <- compute_expected(fs_test)

  fs_col     <- paste0("fs_",     factor_name)
  pctile_col <- paste0("pctile_", factor_name)
  label_col  <- paste0("label_",  factor_name)

  cfa_tbl <- bind_rows(
    tibble(survey = survey, fold_id = fold_id, split = "train",
           row_id = train_df$row_id,
           !!fs_col := fs_train, !!pctile_col := pctile_train,
           !!label_col := label_train) |> bind_cols(exp_train),
    tibble(survey = survey, fold_id = fold_id, split = "test",
           row_id = test_df$row_id,
           !!fs_col := fs_test, !!pctile_col := pctile_test,
           !!label_col := label_test) |> bind_cols(exp_test)
  )

  # CFA parameters (long format)
  std_sol <- standardizedSolution(cfa_fit) |>
    as_tibble() |>
    select(op, lhs, rhs, std.all = est.std)

  cfa_params <- parameterEstimates(cfa_fit) |>
    as_tibble() |>
    left_join(std_sol, by = c("op", "lhs", "rhs")) |>
    filter(op %in% c("=~", "~1")) |>
    mutate(survey = survey, fold_id = fold_id) |>
    select(survey, fold_id, op, lhs, rhs, est, se, z,
           pvalue, std.all)

  # Fit indices
  fit_vals <- fitMeasures(cfa_fit,
                           c("cfi", "tli", "rmsea", "srmr")) |>
    as.list() |> as_tibble()

  omega_val  <- tryCatch(as.numeric(compRelSEM(cfa_fit)),
                          error = function(e) NA_real_)
  ave_val    <- tryCatch(as.numeric(AVE(cfa_fit)),
                          error = function(e) NA_real_)
  factor_var <- lavInspect(cfa_fit, "est")$psi[1, 1]

  cfa_fit_row <- fit_vals |>
    mutate(survey = survey, fold_id = fold_id,
           split = "train",
           omega = omega_val, ave = ave_val,
           factor_var = factor_var) |>
    select(survey, fold_id, split, cfi, tli, rmsea, srmr,
           omega, ave, factor_var)

  # Test-fold CFA baseline (H2: ground truth for imputation
  # vs. deletion comparison). Refit CFA on test data to get
  # complete-data measurement properties for comparison.
  cfa_fit_test <- tryCatch({
    test_cfa <- cfa(cfg$cfa_syntax, data = trust_test,
                     estimator = "ML", meanstructure = TRUE)
    test_fit_vals <- fitMeasures(test_cfa,
                                  c("cfi", "tli", "rmsea",
                                    "srmr")) |>
      as.list() |> as_tibble()
    test_omega <- tryCatch(as.numeric(compRelSEM(test_cfa)),
                            error = function(e) NA_real_)
    test_ave   <- tryCatch(as.numeric(AVE(test_cfa)),
                            error = function(e) NA_real_)
    test_fvar  <- lavInspect(test_cfa, "est")$psi[1, 1]

    test_fit_vals |>
      mutate(survey = survey, fold_id = fold_id,
             split = "test",
             omega = test_omega, ave = test_ave,
             factor_var = test_fvar) |>
      select(survey, fold_id, split, cfi, tli, rmsea, srmr,
             omega, ave, factor_var)
  }, error = function(e) {
    cat(sprintf("    Test-fold CFA failed: %s\n", e$message))
    tibble(survey = survey, fold_id = fold_id,
           split = "test",
           cfi = NA_real_, tli = NA_real_,
           rmsea = NA_real_, srmr = NA_real_,
           omega = NA_real_, ave = NA_real_,
           factor_var = NA_real_)
  })

  # Combine train and test fit rows
  cfa_fit_both <- bind_rows(cfa_fit_row, cfa_fit_test)

  list(cfa_tbl = cfa_tbl, cfa_params = cfa_params,
       cfa_fit = cfa_fit_both)
}
```

# Component 2: Per-Fold IRT/GRM

```{r}
#| label: fn-irt

fit_fold_irt <- function(train_df, test_df, cfg, survey, fold_id) {

  trust_items <- cfg$trust_items
  trust_train <- train_df |>
    select(all_of(trust_items)) |>
    mutate(across(everything(), as.numeric))
  trust_test <- test_df |>
    select(all_of(trust_items)) |>
    mutate(across(everything(), as.numeric))

  item_min <- min(unlist(trust_train), na.rm = TRUE)
  item_max <- max(unlist(trust_train), na.rm = TRUE)

  # Reverse-code for descending scales (LB)
  if (!cfg$trust_ascending) {
    irt_train <- trust_train |>
      mutate(across(everything(), ~ as.integer(item_max + 1L - .)))
    irt_test <- trust_test |>
      mutate(across(everything(), ~ as.integer(item_max + 1L - .)))
  } else {
    irt_train <- trust_train |>
      mutate(across(everything(), as.integer))
    irt_test <- trust_test |>
      mutate(across(everything(), as.integer))
  }

  irt_fit <- mirt(irt_train, 1, itemtype = "graded",
                   verbose = FALSE, SE = TRUE)

  # Discrimination + SE
  item_coefs <- coef(irt_fit, printSE = TRUE)
  irt_params <- map_dfr(trust_items, function(item) {
    mat <- item_coefs[[item]]
    tibble(survey = survey, fold_id = fold_id, item = item,
           a = round(mat["par", "a1"], 3),
           se_a = round(mat["SE", "a1"], 3))
  })

  # Q3 residual correlations
  q3_mat <- residuals(irt_fit, type = "Q3", verbose = FALSE)
  q3_pairs <- map_dfr(seq_along(trust_items), function(i) {
    map_dfr(seq_len(i - 1), function(j) {
      q3_val <- q3_mat[i, j]
      if (abs(q3_val) > Q3_THRESHOLD) {
        tibble(survey = survey, fold_id = fold_id,
               item1 = trust_items[j], item2 = trust_items[i],
               q3 = round(q3_val, 3))
      }
    })
  })

  irt_fit_row <- tibble(
    survey = survey, fold_id = fold_id,
    n_q3_notable = nrow(q3_pairs),
    max_q3 = round(if (nrow(q3_pairs) > 0) {
      max(abs(q3_pairs$q3))
    } else 0, 3)
  )

  # Theta estimates
  theta_train_mat <- fscores(irt_fit, method = "EAP",
                              full.scores.SE = TRUE)
  theta_train <- as.numeric(theta_train_mat[, 1])
  se_train    <- as.numeric(theta_train_mat[, 2])

  theta_test_mat <- fscores(
    irt_fit, response.pattern = as.matrix(irt_test),
    method = "EAP", full.scores.SE = TRUE)
  theta_test <- as.numeric(theta_test_mat[, 1])
  se_test    <- as.numeric(theta_test_mat[, 2])

  ecdf_fn      <- ecdf(theta_train)
  pctile_train <- round(ecdf_fn(theta_train) * 100)
  pctile_test  <- round(ecdf_fn(theta_test)  * 100)
  label_train  <- label_trust(theta_train, theta_train)
  label_test   <- label_trust(theta_test,  theta_train)

  irt_tbl <- bind_rows(
    tibble(survey = survey, fold_id = fold_id, split = "train",
           row_id = train_df$row_id, irt_theta = theta_train,
           irt_se = round(se_train, 3),
           irt_pctile = pctile_train, irt_label = label_train),
    tibble(survey = survey, fold_id = fold_id, split = "test",
           row_id = test_df$row_id, irt_theta = theta_test,
           irt_se = round(se_test, 3),
           irt_pctile = pctile_test, irt_label = label_test)
  )

  list(irt_tbl = irt_tbl, irt_params = irt_params,
       irt_fit = irt_fit_row, q3_pairs = q3_pairs,
       irt_model = irt_fit)
}
```

# Component 3: Demographic Profiling

This component replaces MI, DIF, and theta regression with a
data-driven approach that produces information LLMs can directly
use. For each demographic variable, we compute:

1. **Screening:** Eta-squared with both IRT theta and CFA factor
   score. Demographics must pass the threshold (eta² ≥
   `r ETA_SQ_THRESHOLD`) for at least one latent variable to be
   included in prompts for that survey. Both are reported so the
   reader can see whether CFA and IRT screening results diverge.
2. **Group means (Layer 1):** Mean trust rating per item per
   demographic group. Provides direction and magnitude.
3. **Model-based probabilities (Layer 2):** For each demographic
   group's mean theta, compute $P(X_j \geq k \mid \bar{\theta}_g)$
   using the IRT model. This transforms psychometric parameters
   into probability statements the LLM can interpret.
4. **Conditional distributions (Layer 3):** Empirical
   $P(X_j = k \mid \text{group})$ for the strongest associations.

```{r}
#| label: fn-demo-profile

fit_fold_demo_profile <- function(train_df, irt_train_tbl,
                                   cfa_train_tbl, irt_model,
                                   cfg, survey, fold_id) {

  trust_items <- cfg$trust_items
  demo_vars   <- cfg$demo_vars
  factor_name <- cfg$factor_names
  fs_col      <- paste0("fs_", factor_name)

  item_min    <- min(as.numeric(unlist(
    train_df |> select(all_of(trust_items))
  )), na.rm = TRUE)
  item_max    <- max(as.numeric(unlist(
    train_df |> select(all_of(trust_items))
  )), na.rm = TRUE)
  n_cats      <- item_max - item_min + 1L

  # Merge both theta and CFA factor score onto training data
  merged <- train_df |>
    left_join(irt_train_tbl |> select(row_id, irt_theta),
              by = "row_id") |>
    left_join(cfa_train_tbl |> select(row_id, cfa_fs = !!sym(fs_col)),
              by = "row_id")

  ecdf_theta <- ecdf(merged$irt_theta)

  # ── Step 1: Screening ──────────────────────────────────────
  # Eta-squared of each demographic with BOTH IRT theta and
  # CFA factor score. A demographic passes if it meets the
  # threshold for at least one latent variable.
  screening <- map_dfr(demo_vars, function(dv) {
    if (!dv %in% names(merged)) {
      return(tibble(survey = survey, fold_id = fold_id,
                    demo_var = dv,
                    eta_sq_theta = NA_real_,
                    eta_sq_cfa = NA_real_,
                    F_theta = NA_real_, p_theta = NA_real_,
                    F_cfa = NA_real_, p_cfa = NA_real_,
                    n = 0L, passes = FALSE))
    }

    # IRT theta screening
    irt_result <- compute_eta_sq(merged$irt_theta, merged[[dv]])

    # CFA factor score screening
    cfa_result <- compute_eta_sq(merged$cfa_fs, merged[[dv]])

    tibble(
      survey       = survey,
      fold_id      = fold_id,
      demo_var     = dv,
      eta_sq_theta = irt_result$eta_sq,
      eta_sq_cfa   = cfa_result$eta_sq,
      F_theta      = irt_result$F_stat,
      p_theta      = irt_result$p_value,
      F_cfa        = cfa_result$F_stat,
      p_cfa        = cfa_result$p_value,
      n            = irt_result$n,
      # Passes if EITHER latent variable meets threshold
      passes       = (!is.na(irt_result$eta_sq) &
                        irt_result$eta_sq >= ETA_SQ_THRESHOLD) |
                     (!is.na(cfa_result$eta_sq) &
                        cfa_result$eta_sq >= ETA_SQ_THRESHOLD)
    )
  })

  # Demographics that pass screening for this fold
  demo_passed <- screening |>
    filter(passes == TRUE) |>
    pull(demo_var)

  cat(sprintf("    Demo screening: %s pass → [%s]\n",
              length(demo_passed),
              str_c(demo_passed, collapse = ", ")))

  # ── Step 2: Group means per item (Layer 1) ─────────────────
  # Only for demographics that pass screening
  group_means_items <- map_dfr(demo_passed, function(dv) {
    map_dfr(trust_items, function(item) {
      merged |>
        filter(!is.na(.data[[dv]]), !is.na(.data[[item]])) |>
        group_by(group = as.character(.data[[dv]])) |>
        summarise(
          mean_trust = round(mean(as.numeric(.data[[item]])), 3),
          sd_trust   = round(sd(as.numeric(.data[[item]])), 3),
          n          = n(),
          .groups    = "drop"
        ) |>
        mutate(survey = survey, fold_id = fold_id,
               demo_var = dv, item = item, .before = 1)
    })
  })

  # ── Step 3: Model-based probabilities (Layer 2) ────────────
  # For each demographic group, compute mean theta and then use
  # the IRT model to get P(X_j >= k | mean_theta_group)
  # This transforms IRT parameters into probability statements
  group_means_theta <- map_dfr(demo_passed, function(dv) {
    merged |>
      filter(!is.na(.data[[dv]]), !is.na(irt_theta)) |>
      group_by(group = as.character(.data[[dv]])) |>
      summarise(
        mean_theta = mean(irt_theta),
        sd_theta   = sd(irt_theta),
        n          = n(),
        .groups    = "drop"
      ) |>
      mutate(
        pctile = round(ecdf_theta(mean_theta) * 100),
        survey = survey, fold_id = fold_id,
        demo_var = dv, .before = 1
      )
  })

  # Model-based conditional probabilities:
  # P(X_j >= k | group mean theta) for each item × group
  model_probs <- map_dfr(demo_passed, function(dv) {

    group_thetas <- group_means_theta |>
      filter(demo_var == dv) |>
      mutate(group = as.character(group)) |>
      select(group, mean_theta)

    pmap_dfr(group_thetas, function(group, mean_theta) {
      map_dfr(trust_items, function(item) {

        item_obj  <- extract.item(irt_model, item)
        raw_probs <- as.numeric(probtrace(item_obj, mean_theta))

        # Reverse probabilities for descending scales
        probs <- if (!cfg$trust_ascending) {
          rev(raw_probs)
        } else {
          raw_probs
        }

        # Category labels (original scale)
        cats <- seq(item_min, item_max)

        # Cumulative P(X >= k) for each category
        cum_probs <- map_dbl(seq_along(cats), function(k) {
          sum(probs[k:length(probs)])
        })

        # Find midpoint category
        midpoint <- ceiling(n_cats / 2) + item_min - 1L

        # P(higher trust) — scale-direction-aware
        # For ascending scales: higher trust = higher categories
        # For descending scales (after rev): higher trust = lower
        # category numbers on the original scale
        p_high_trust <- if (cfg$trust_ascending) {
          sum(probs[which(cats > midpoint)])
        } else {
          sum(probs[which(cats < midpoint)])
        }

        tibble(
          survey    = survey,
          fold_id   = fold_id,
          demo_var  = dv,
          group     = group,
          item      = item,
          modal_cat = cats[which.max(probs)],
          p_high_trust = round(p_high_trust, 3),
          p_at_modal  = round(max(probs), 3),
          prob_spread = list(
            tibble(category = cats, prob = round(probs, 4),
                   p_geq = round(cum_probs, 4))
          )
        )
      })
    })
  })

  # ── Step 4: Conditional distributions (Layer 3) ────────────
  # Empirical P(X_j = k | group) for all items × passed demos
  conditional_dists <- map_dfr(demo_passed, function(dv) {
    map_dfr(trust_items, function(item) {
      merged |>
        filter(!is.na(.data[[dv]]),
               !is.na(.data[[item]])) |>
        count(group = as.character(.data[[dv]]),
              category = as.integer(.data[[item]])) |>
        group_by(group) |>
        mutate(prop = round(n / sum(n), 4)) |>
        ungroup() |>
        mutate(survey = survey, fold_id = fold_id,
               demo_var = dv, item = item, .before = 1)
    })
  })

  list(
    screening         = screening,
    group_means_items = group_means_items,
    group_means_theta = group_means_theta,
    model_probs       = model_probs,
    conditional_dists = conditional_dists
  )
}
```

# Main Fitting Loop

```{r}
#| label: fit-all
#| results: hide

fit_all_folds <- function(fold_df, cfg, survey_name) {

  fold_ids <- sort(unique(fold_df$fold_id))

  results <- map(fold_ids, function(fi) {

    cat("\n== Fitting:", survey_name, "| Fold", fi, "==\n")

    train_df <- fold_df |>
      filter(fold_id == fi, split == "train") |>
      arrange(row_id)
    test_df <- fold_df |>
      filter(fold_id == fi, split == "test") |>
      arrange(row_id)

    cat("  CFA...\n")
    cfa_out <- fit_fold_cfa(train_df, test_df, cfg,
                             survey_name, fi)

    cat("  IRT/GRM...\n")
    irt_out <- fit_fold_irt(train_df, test_df, cfg,
                             survey_name, fi)

    cat("  Demographic profiling...\n")
    irt_train <- irt_out$irt_tbl |> filter(split == "train")
    cfa_train <- cfa_out$cfa_tbl |> filter(split == "train")
    demo_out <- fit_fold_demo_profile(
      train_df, irt_train, cfa_train, irt_out$irt_model,
      cfg, survey_name, fi)

    cat("  Done.\n")

    list(cfa = cfa_out, irt = irt_out, demo = demo_out)
  })

  # ── Assemble all outputs ──────────────────────────────────
  list(
    cfa_tbl      = map_dfr(results, function(x) x$cfa$cfa_tbl),
    cfa_params   = map_dfr(results, function(x) x$cfa$cfa_params),
    cfa_fit      = map_dfr(results, function(x) x$cfa$cfa_fit),
    irt_tbl      = map_dfr(results, function(x) x$irt$irt_tbl),
    irt_params   = map_dfr(results, function(x) x$irt$irt_params),
    irt_fit      = map_dfr(results, function(x) x$irt$irt_fit),
    q3_pairs     = map_dfr(results, function(x) x$irt$q3_pairs),
    irt_models   = map(results, function(x) x$irt$irt_model) |>
                     set_names(fold_ids),
    # Demographic profiling
    demo_screening    = map_dfr(results,
                                function(x) x$demo$screening),
    group_means_items = map_dfr(results,
                                function(x) x$demo$group_means_items),
    group_means_theta = map_dfr(results,
                                function(x) x$demo$group_means_theta),
    model_probs       = map_dfr(results,
                                function(x) x$demo$model_probs),
    conditional_dists = map_dfr(results,
                                function(x) x$demo$conditional_dists)
  )
}

lapop_results <- fit_all_folds(folds_list$lapop, config$lapop,
                                "lapop")
lb_results    <- fit_all_folds(folds_list$lb, config$lb, "lb")
```

# Assemble Output Objects

```{r}
#| label: assemble

# CFA
cfa_list       <- list(lapop = lapop_results$cfa_tbl,
                         lb = lb_results$cfa_tbl)
cfa_params_tbl <- bind_rows(lapop_results$cfa_params,
                              lb_results$cfa_params)
cfa_fit_tbl    <- bind_rows(lapop_results$cfa_fit,
                              lb_results$cfa_fit)

# IRT
irt_list       <- list(lapop = lapop_results$irt_tbl,
                         lb = lb_results$irt_tbl)
irt_params_tbl <- bind_rows(lapop_results$irt_params,
                              lb_results$irt_params)
irt_fit_tbl    <- bind_rows(lapop_results$irt_fit,
                              lb_results$irt_fit)
q3_pairs_tbl   <- bind_rows(lapop_results$q3_pairs,
                              lb_results$q3_pairs)
irt_models     <- list(lapop = lapop_results$irt_models,
                         lb = lb_results$irt_models)

# Demographic profiling
demo_screening_tbl    <- bind_rows(lapop_results$demo_screening,
                                    lb_results$demo_screening)
group_means_items_tbl <- bind_rows(
  lapop_results$group_means_items,
  lb_results$group_means_items)
group_means_theta_tbl <- bind_rows(
  lapop_results$group_means_theta,
  lb_results$group_means_theta)
model_probs_tbl       <- bind_rows(lapop_results$model_probs,
                                    lb_results$model_probs)
conditional_dists_tbl <- bind_rows(
  lapop_results$conditional_dists,
  lb_results$conditional_dists)

# Assemble the demo_profile_tbl as a list of the three layers
# plus model probabilities
demo_profile_tbl <- list(
  screening         = demo_screening_tbl,
  group_means_items = group_means_items_tbl,
  group_means_theta = group_means_theta_tbl,
  model_probs       = model_probs_tbl,
  conditional_dists = conditional_dists_tbl
)

cat("Assembly complete.\n")
```

# Validation

```{r}
#| label: validation

bind_rows(
  tibble(Survey = "lapop",
         CFA_rows = nrow(cfa_list$lapop),
         IRT_rows = nrow(irt_list$lapop),
         Fold_rows = nrow(folds_list$lapop),
         CFA_ok = nrow(cfa_list$lapop) == nrow(folds_list$lapop),
         IRT_ok = nrow(irt_list$lapop) == nrow(folds_list$lapop)),
  tibble(Survey = "lb",
         CFA_rows = nrow(cfa_list$lb),
         IRT_rows = nrow(irt_list$lb),
         Fold_rows = nrow(folds_list$lb),
         CFA_ok = nrow(cfa_list$lb) == nrow(folds_list$lb),
         IRT_ok = nrow(irt_list$lb) == nrow(folds_list$lb))
) |> knitr::kable(align = "lccccl")
```

# CFA Fit Summary

```{r}
#| label: tbl-cfa-fit
#| tbl-cap: "Mean CFA Fit Across 5 Folds (Train = model estimation, Test = H2 baseline)"

cfa_fit_tbl |>
  group_by(survey, split) |>
  summarise(across(c(cfi, tli, rmsea, srmr, omega, ave),
                    mean, na.rm = TRUE), .groups = "drop") |>
  mutate(across(where(is.numeric), function(x) round(x, 3))) |>
  knitr::kable(col.names = c("Survey", "Split", "CFI", "TLI",
                              "RMSEA", "SRMR", "ω", "AVE"),
               align = "llcccccc")
```

# IRT Discrimination Summary

```{r}
#| label: tbl-irt
#| tbl-cap: "Mean Item Discrimination Across 5 Folds"

irt_params_tbl |>
  group_by(survey, item) |>
  summarise(mean_a = round(mean(a), 3),
            sd_a = round(sd(a), 3), .groups = "drop") |>
  arrange(survey, desc(mean_a)) |>
  knitr::kable(col.names = c("Survey", "Item", "Mean a",
                              "SD(a)"),
               align = "llcc")
```

# Demographic Screening Results

```{r}
#| label: tbl-demo-screening
#| tbl-cap: "Demographic Screening: Eta² with IRT Theta and CFA Factor Score (Averaged Across 5 Folds)"

demo_screening_tbl |>
  group_by(survey, demo_var) |>
  summarise(
    mean_eta_theta = round(mean(eta_sq_theta, na.rm = TRUE), 4),
    mean_eta_cfa   = round(mean(eta_sq_cfa, na.rm = TRUE), 4),
    n_folds_pass   = sum(passes),
    .groups = "drop"
  ) |>
  mutate(
    max_eta = pmax(mean_eta_theta, mean_eta_cfa, na.rm = TRUE),
    verdict = case_when(
      max_eta >= 0.01 ~ "KEEP",
      max_eta >= ETA_SQ_THRESHOLD ~ "MARGINAL-KEEP",
      TRUE ~ "DROP"
    )
  ) |>
  arrange(survey, desc(max_eta)) |>
  select(-max_eta) |>
  knitr::kable(
    col.names = c("Survey", "Demographic", "η²(θ)",
                  "η²(CFA)", "Folds Pass", "Verdict"),
    align = "llcccc"
  )
```

# Demographic Group Means (Theta)

```{r}
#| label: tbl-group-theta
#| tbl-cap: "Mean Theta by Demographic Group (Averaged Across 5 Folds, Surviving Demographics Only)"

group_means_theta_tbl |>
  group_by(survey, demo_var, group) |>
  summarise(
    mean_theta = round(mean(mean_theta), 3),
    mean_pctile = round(mean(pctile)),
    mean_n = round(mean(n)),
    .groups = "drop"
  ) |>
  arrange(survey, demo_var, desc(mean_theta)) |>
  knitr::kable(
    col.names = c("Survey", "Demographic", "Group",
                  "Mean θ", "Percentile", "N"),
    align = "lllccc"
  )
```

# Model-Based Probability Context Example

```{r}
#| label: tbl-model-probs-example
#| tbl-cap: "IRT Model-Based P(high trust) by Demographic Group — Fold 1, Top Items"

# Show the top 5 items by between-group probability range
# for each survey (fold 1 only for display)
model_probs_tbl |>
  filter(fold_id == 1) |>
  group_by(survey, demo_var, item) |>
  summarise(
    range_p = max(p_high_trust) - min(p_high_trust),
    .groups = "drop"
  ) |>
  group_by(survey) |>
  slice_max(range_p, n = 6) |>
  ungroup() |>
  left_join(
    model_probs_tbl |>
      filter(fold_id == 1) |>
      select(survey, demo_var, item, group,
             p_high_trust, modal_cat),
    by = c("survey", "demo_var", "item")
  ) |>
  select(survey, demo_var, item, group,
         p_high_trust, modal_cat) |>
  arrange(survey, demo_var, item,
          desc(p_high_trust)) |>
  knitr::kable(
    col.names = c("Survey", "Demo", "Item", "Group",
                  "P(high trust)", "Modal Cat"),
    align = "lllccc"
  )
```

\newpage

# Contour Plots: Theta Distribution by Demographic Group

```{r}
#| label: fig-theta-ridgeline
#| fig-cap: "IRT Theta Distribution by Demographic Group. Only demographics that passed screening are shown. Width of ridges reflects the density of respondents at each theta level."
#| fig-height: 12
#| fig-width: 10

# Build combined data for ridgeline plots
# Use fold 1 training data for display
ridge_data <- map_dfr(c("lapop", "lb"), function(survey_name) {

  # Get surviving demographics for this survey
  # A demographic survives if max(eta_theta, eta_cfa) >= threshold
  survivors <- demo_screening_tbl |>
    filter(survey == survey_name) |>
    group_by(demo_var) |>
    summarise(
      mean_eta_theta = mean(eta_sq_theta, na.rm = TRUE),
      mean_eta_cfa   = mean(eta_sq_cfa, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(max_eta = pmax(mean_eta_theta, mean_eta_cfa,
                          na.rm = TRUE)) |>
    filter(max_eta >= ETA_SQ_THRESHOLD) |>
    pull(demo_var)

  # Get fold 1 training data with theta
  train_df <- folds_list[[survey_name]] |>
    filter(fold_id == 1, split == "train")

  theta_df <- irt_list[[survey_name]] |>
    filter(fold_id == 1, split == "train") |>
    select(row_id, irt_theta)

  merged <- train_df |>
    left_join(theta_df, by = "row_id")

  # Pivot demographics to long format
  map_dfr(survivors, function(dv) {
    merged |>
      filter(!is.na(.data[[dv]]), !is.na(irt_theta)) |>
      transmute(
        survey = if_else(survey_name == "lapop",
                         "LAPOP Mexico (7-point)",
                         "Latinobarómetro Colombia (4-point)"),
        demo_var = dv,
        group = as.character(.data[[dv]]),
        irt_theta = irt_theta
      )
  })
})

# Ridgeline plot
ggplot(ridge_data,
       aes(x = irt_theta, y = group, fill = group)) +
  geom_density_ridges(
    alpha = 0.7, scale = 1.2,
    rel_min_height = 0.01
  ) +
  facet_grid(demo_var ~ survey, scales = "free_y",
             space = "free_y") +
  scale_fill_viridis_d(option = "D", end = 0.85) +
  labs(
    x = expression(IRT ~ theta ~ (latent ~ trust)),
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    strip.text.y = element_text(angle = 0, face = "bold"),
    strip.text.x = element_text(face = "bold")
  )
```

# Generate Masking Table

Each test respondent is randomly assigned 1--4 missing items, with
the specific masked items selected at random.

```{r}
#| label: masking-table

masking_tbl <- map_dfr(names(config), function(survey_name) {
  items <- config[[survey_name]]$trust_items

  map_dfr(1:N_FOLDS, function(fold_i) {
    test_ids <- folds_list[[survey_name]] |>
      filter(fold_id == fold_i, split == "test") |>
      pull(row_id)

    n_test   <- length(test_ids)
    n_masked <- sample(1:4, n_test, replace = TRUE)

    map2_dfr(test_ids, n_masked, function(rid, k) {
      tibble(
        survey      = survey_name,
        fold_id     = fold_i,
        row_id      = rid,
        n_masked    = k,
        masked_item = list(sample(items, size = k))
      )
    })
  })
})

cat("Masking table:", nrow(masking_tbl), "rows\n\n")
masking_tbl |>
  count(survey, n_masked) |>
  knitr::kable(caption = "Masking Distribution",
               align = "lcc")
```

# Save Outputs

```{r}
#| label: save-outputs

# Primary fold data
output_path <- file.path(results_dir, "fold_data.RData")
save(folds_list, cfa_list, irt_list, masking_tbl,
     file = output_path)
cat("Saved fold_data.RData:",
    round(file.size(output_path) / 1e6, 2), "MB\n")

# Also save masking table as standalone RDS
saveRDS(masking_tbl, file.path(results_dir,
                                "masking_table.rds"))
cat("Saved masking_table.rds\n")

# Psychometric reporting tables
reporting_path <- file.path(results_dir,
                             "cfa_reporting.RData")
save(cfa_params_tbl, cfa_fit_tbl,
     irt_params_tbl, irt_fit_tbl, q3_pairs_tbl,
     demo_profile_tbl,
     file = reporting_path)
cat("Saved cfa_reporting.RData:",
    round(file.size(reporting_path) / 1e6, 2), "MB\n")

# IRT model objects (needed for imputation scripts)
irt_path <- file.path(results_dir, "irt_models.RData")
save(irt_models, file = irt_path)
cat("Saved irt_models.RData:",
    round(file.size(irt_path) / 1e6, 2), "MB\n")

# Summary of saved objects
tibble(
  Object = c("folds_list$lapop", "folds_list$lb",
              "cfa_list$lapop", "cfa_list$lb",
              "irt_list$lapop", "irt_list$lb",
              "cfa_params_tbl", "cfa_fit_tbl",
              "irt_params_tbl", "irt_fit_tbl",
              "q3_pairs_tbl",
              "demo_profile_tbl$screening",
              "demo_profile_tbl$group_means_items",
              "demo_profile_tbl$group_means_theta",
              "demo_profile_tbl$model_probs",
              "demo_profile_tbl$conditional_dists",
              "masking_tbl",
              "irt_models$lapop", "irt_models$lb"),
  Rows = c(nrow(folds_list$lapop), nrow(folds_list$lb),
            nrow(cfa_list$lapop), nrow(cfa_list$lb),
            nrow(irt_list$lapop), nrow(irt_list$lb),
            nrow(cfa_params_tbl), nrow(cfa_fit_tbl),
            nrow(irt_params_tbl), nrow(irt_fit_tbl),
            nrow(q3_pairs_tbl),
            nrow(demo_profile_tbl$screening),
            nrow(demo_profile_tbl$group_means_items),
            nrow(demo_profile_tbl$group_means_theta),
            nrow(demo_profile_tbl$model_probs),
            nrow(demo_profile_tbl$conditional_dists),
            nrow(masking_tbl),
            length(irt_models$lapop),
            length(irt_models$lb)),
  Saved_In = c(rep("fold_data", 6),
                rep("cfa_reporting", 10),
                "fold_data",
                rep("irt_models", 2))
) |> knitr::kable(align = "lcc")
```
