---
title: "L2L Framework: TSE Evaluation"
subtitle: "Script 05 — RQ1 (Item Selection), RQ2 (Construct Recovery), RQ3 (Item Measurement Error)"
author: "Kevin Linares"
date: today
format:
  pdf:
    colorlinks: true
    fig-pos: "H"
    geometry: margin=1in
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  echo: true
  warning: false
  message: false
  cache: false
editor:
  markdown:
    wrap: 72
---

# Overview

Three CFA/IRT model fits per fold define the evaluation:

- **Gold standard:** CFA/IRT fit to full fold (training + test,
  no masking).
- **Listwise deletion:** CFA/IRT fit to training data only.
- **Imputed:** CFA/IRT fit to training + imputed test.

## Research Questions

- **RQ1 — Psychometric context utilization:** Select best
  LLM-tier on item-level metrics (weighted kappa, accuracy).
- **RQ2 — Construct-level recovery:** Compare gold, deletion,
  best L2L, best traditional on rank preservation (r) and
  scale fidelity (regression slope).
- **RQ3 — Item measurement error:** Quantify how imputation
  distorts IRT item parameters (discrimination, thresholds)
  relative to gold standard, using Biemer decomposition of
  item parameter MSE into systematic (Bias²) and random
  (Variance) components. Counterfactual = listwise deletion.

# Setup

```{r}
#| label: setup

library(lavaan)
library(mirt)
library(semTools)
library(viridis)
library(tictoc)

set.seed(721)

results_dir <- "D:/repos/UMD_classes_code/TSE_II_SURV721/results"

llm_results  <- readRDS(file.path(results_dir,
                                   "results_item.rds"))
trad_results <- readRDS(file.path(results_dir,
                                   "results_trad_item.rds"))
load(file.path(results_dir, "fold_data.RData"))
load(file.path(results_dir, "cfa_reporting.RData"))
load(file.path(results_dir, "irt_models.RData"))
masking_tbl <- readRDS(file.path(results_dir,
                                  "masking_table.rds"))

# ── tidyverse loaded last ────────────────────────────────────
library(tidyverse)
```

# Data Preparation

```{r}
#| label: data-prep

shared_cols <- c("survey", "fold_id", "n_masked", "tier",
                 "method", "row_id", "item", "actual",
                 "predicted", "cfa_baseline", "irt_baseline")

all_results <- bind_rows(
  trad_results |>
    filter(!is.na(predicted)) |>
    select(all_of(shared_cols)) |>
    mutate(method_type = "Traditional"),
  llm_results |>
    filter(!is.na(predicted)) |>
    select(all_of(shared_cols)) |>
    mutate(method_type = "LLM")
) |>
  filter(method != "gemma3_27b") |>
  mutate(
    method_label = recode(method,
      llama4_maverick = "Maverick",
      claude_sonnet   = "Claude Sonnet",
      mixgb           = "mixgb",
      miceRanger      = "miceRanger"),
    method_type = factor(method_type,
                         levels = c("Traditional", "LLM")),
    tier = factor(tier, levels = c("S", "P1", "P2")),
    survey_label = if_else(survey == "lapop",
                           "LAPOP Mexico",
                           "Latinobarómetro Colombia")
  )

survey_config <- list(
  lapop = list(
    trust_items = c(
      "justice_system", "electoral_tribunal", "armed_forces",
      "legislature", "public_ministry", "police", "auditor",
      "political_parties", "supreme_court", "municipality",
      "media", "elections"),
    cfa_syntax = '
      Trust =~ justice_system + electoral_tribunal +
               armed_forces + legislature + public_ministry +
               police + auditor + political_parties +
               supreme_court + municipality + media + elections
      armed_forces ~~ legislature
      public_ministry ~~ auditor
    ',
    trust_ascending = TRUE),
  lb = list(
    trust_items = c(
      "Congress", "Natl_Govt", "Judiciary", "Parties",
      "Electoral_Inst", "President"),
    cfa_syntax = '
      Political =~ Congress + Natl_Govt + Judiciary + Parties +
                   Electoral_Inst + President
      Natl_Govt ~~ President
    ',
    trust_ascending = FALSE)
)

cat("Combined (no Gemma):", nrow(all_results), "rows\n")
```

# Helper Functions

```{r}
#| label: fn-helpers

# ── CFA: fit and score ───────────────────────────────────────
fit_cfa_score <- function(data_mat, cfa_syntax, row_ids) {
  tryCatch({
    fit <- cfa(cfa_syntax, data = data_mat,
               estimator = "ML", meanstructure = TRUE)
    fs <- as.numeric(lavPredict(fit, method = "regression"))
    omega <- as.numeric(semTools::compRelSEM(fit))
    list(scores = tibble(row_id = row_ids, fs = fs),
         omega = omega)
  }, error = function(e) {
    cat("  CFA failed:", conditionMessage(e), "\n")
    list(scores = tibble(row_id = row_ids,
                         fs = rep(NA_real_, length(row_ids))),
         omega = NA_real_)
  })
}

# ── IRT: fit and score (reverse-codes LB) ────────────────────
fit_irt_score <- function(data_mat, trust_ascending, row_ids) {
  item_max <- max(data_mat, na.rm = TRUE)
  irt_data <- if (!trust_ascending) {
    data_mat |>
      mutate(across(everything(),
                    function(x) as.integer(item_max + 1L - x)))
  } else {
    data_mat |>
      mutate(across(everything(), as.integer))
  }
  tryCatch({
    fit <- mirt(irt_data, 1, itemtype = "graded",
                verbose = FALSE)
    theta <- as.numeric(fscores(fit, method = "EAP")[, 1])
    list(scores = tibble(row_id = row_ids, theta = theta))
  }, error = function(e) {
    cat("  IRT failed:", conditionMessage(e), "\n")
    list(scores = tibble(row_id = row_ids,
                         theta = rep(NA_real_,
                                     length(row_ids))))
  })
}

# ── Combined fit: CFA + IRT + correlation matrix ─────────────
fit_both_models <- function(data_df, cfg, label = "") {
  trust_items <- cfg$trust_items
  row_ids <- data_df$row_id
  data_mat <- data_df |>
    select(all_of(trust_items)) |>
    mutate(across(everything(), as.numeric))
  cat("  Fitting:", label, "| N =", nrow(data_mat), "\n")
  cfa_out <- fit_cfa_score(data_mat, cfg$cfa_syntax, row_ids)
  irt_out <- fit_irt_score(data_mat, cfg$trust_ascending,
                            row_ids)
  cor_mat <- cor(data_mat, use = "pairwise.complete.obs")
  scores <- cfa_out$scores |>
    left_join(irt_out$scores, by = "row_id")
  list(scores = scores, omega = cfa_out$omega,
       cor_mat = cor_mat,
       mean_theta = mean(scores$theta, na.rm = TRUE),
       sd_theta = sd(scores$theta, na.rm = TRUE),
       n = nrow(scores))
}

# ── Reconstruct training + imputed test ──────────────────────
reconstruct_full_data <- function(survey_name, fold_i,
                                   method_name, tier_name) {
  cfg <- survey_config[[survey_name]]
  train_df <- folds_list[[survey_name]] |>
    filter(fold_id == fold_i, split == "train") |>
    select(row_id, all_of(cfg$trust_items)) |>
    mutate(across(all_of(cfg$trust_items), as.numeric))
  test_df <- folds_list[[survey_name]] |>
    filter(fold_id == fold_i, split == "test") |>
    select(row_id, all_of(cfg$trust_items)) |>
    mutate(across(all_of(cfg$trust_items), as.numeric))
  preds <- all_results |>
    filter(survey == survey_name, fold_id == fold_i,
           method == method_name, tier == tier_name) |>
    select(row_id, item, predicted) |>
    mutate(predicted = as.numeric(predicted))
  if (nrow(preds) == 0L) return(NULL)
  preds_wide <- preds |>
    pivot_wider(names_from = item, values_from = predicted,
                names_prefix = "imp_")
  test_merged <- test_df |>
    left_join(preds_wide, by = "row_id")
  test_imputed <- map_dfc(cfg$trust_items, function(it) {
    actual_val <- test_merged[[it]]
    imp_col <- paste0("imp_", it)
    imp_val <- if (imp_col %in% names(test_merged)) {
      test_merged[[imp_col]]
    } else { NA_real_ }
    tibble(!!it := if_else(!is.na(imp_val),
                           imp_val, actual_val))
  }) |>
    mutate(row_id = test_merged$row_id)
  bind_rows(train_df, test_imputed)
}

# ── Weighted kappa (quadratic weights) ───────────────────────
compute_weighted_kappa <- function(actual, predicted) {
  cats <- sort(unique(c(actual, predicted)))
  k <- length(cats)
  if (k < 2) return(NA_real_)
  obs_tbl <- table(factor(actual, levels = cats),
                   factor(predicted, levels = cats))
  n <- sum(obs_tbl)
  row_m <- rowSums(obs_tbl)
  col_m <- colSums(obs_tbl)
  exp_tbl <- outer(row_m, col_m) / n
  w <- outer(seq_along(cats), seq_along(cats),
             function(i, j) (i - j)^2 / (k - 1)^2)
  1 - sum(w * obs_tbl) / sum(w * exp_tbl)
}

# ── IRT: fit and extract item parameters ─────────────────────
# Returns discrimination (a) and raw intercepts (d) per item
# Uses coef(fit, printSE = TRUE) pattern from Script 02
fit_irt_extract <- function(data_df, cfg) {
  items <- cfg$trust_items
  data_mat <- data_df |>
    select(all_of(items)) |>
    mutate(across(everything(), as.numeric))
  item_max <- max(data_mat, na.rm = TRUE)
  irt_data <- if (!cfg$trust_ascending) {
    data_mat |>
      mutate(across(everything(),
                    function(x) as.integer(item_max + 1L - x)))
  } else {
    data_mat |>
      mutate(across(everything(), as.integer))
  }

  fit <- mirt(irt_data, 1, itemtype = "graded",
              verbose = FALSE)
  item_coefs <- coef(fit, printSE = TRUE)

  map_dfr(items, function(item_name) {
    mat <- item_coefs[[item_name]]
    par_row <- mat["par", ]
    param_names <- names(par_row)

    # Discrimination
    a_val <- par_row[["a1"]]

    # Raw intercepts (d) — NOT converted to b
    d_cols <- param_names[grepl("^d", param_names)]

    # Also compute b = -d/a for reference
    bind_rows(
      tibble(item = item_name, param = "a", value = a_val),
      map_dfr(d_cols, function(dc) {
        d_val <- par_row[[dc]]
        b_label <- gsub("^d", "b", dc)
        tibble(item = item_name, param = dc, value = d_val)
      })
    )
  })
}
```

# Model Fitting

```{r}
#| label: gold-standard

tic("Gold standard fits")
gold_fits <- map_dfr(names(survey_config), function(s) {
  cfg <- survey_config[[s]]
  map_dfr(1:5, function(fi) {
    full_df <- folds_list[[s]] |>
      filter(fold_id == fi) |>
      select(row_id, all_of(cfg$trust_items)) |>
      mutate(across(all_of(cfg$trust_items), as.numeric))
    result <- fit_both_models(
      full_df, cfg, paste("Gold", s, "fold", fi))
    tibble(survey = s, fold_id = fi,
           condition = "Gold",
           method_label = "Gold Standard",
           omega = result$omega,
           mean_theta = result$mean_theta,
           sd_theta = result$sd_theta,
           n = result$n,
           scores = list(result$scores),
           cor_mat = list(result$cor_mat))
  })
})
toc()
```

```{r}
#| label: deletion

tic("Deletion fits")
deletion_fits <- map_dfr(names(survey_config), function(s) {
  cfg <- survey_config[[s]]
  map_dfr(1:5, function(fi) {
    train_df <- folds_list[[s]] |>
      filter(fold_id == fi, split == "train") |>
      select(row_id, all_of(cfg$trust_items)) |>
      mutate(across(all_of(cfg$trust_items), as.numeric))
    result <- fit_both_models(
      train_df, cfg, paste("Deletion", s, "fold", fi))
    tibble(survey = s, fold_id = fi,
           condition = "Deletion",
           method_label = "Listwise Deletion",
           omega = result$omega,
           mean_theta = result$mean_theta,
           sd_theta = result$sd_theta,
           n = result$n,
           scores = list(result$scores),
           cor_mat = list(result$cor_mat))
  })
})
toc()
```

```{r}
#| label: imputed-fits

tic("All imputed fits")
method_tier_grid <- all_results |>
  distinct(survey, method, method_label, method_type, tier)
cat("Fitting", nrow(method_tier_grid),
    "method-tier combos × 5 folds\n")

imputed_fits <- pmap_dfr(method_tier_grid, function(
    survey, method, method_label, method_type, tier) {
  cfg <- survey_config[[survey]]
  map_dfr(1:5, function(fi) {
    recon_df <- reconstruct_full_data(
      survey, fi, method, as.character(tier))
    if (is.null(recon_df)) {
      return(tibble(
        survey = survey, fold_id = fi,
        condition = "Imputed",
        method = method, method_label = method_label,
        method_type = as.character(method_type),
        tier = as.character(tier),
        omega = NA_real_,
        mean_theta = NA_real_, sd_theta = NA_real_,
        n = 0L, scores = list(NULL),
        cor_mat = list(NULL)))
    }
    result <- fit_both_models(
      recon_df, cfg,
      paste(method_label, tier, survey, "fold", fi))
    tibble(survey = survey, fold_id = fi,
           condition = "Imputed",
           method = method, method_label = method_label,
           method_type = as.character(method_type),
           tier = as.character(tier),
           omega = result$omega,
           mean_theta = result$mean_theta,
           sd_theta = result$sd_theta,
           n = result$n,
           scores = list(result$scores),
           cor_mat = list(result$cor_mat))
  })
})
toc()
```

\newpage

# RQ1: Psychometric Context Utilization

## Item-Level Metrics

```{r}
#| label: rq1-items

rq1_item <- all_results |>
  group_by(survey, survey_label, method, method_label,
           method_type, tier, fold_id) |>
  summarise(
    accuracy = mean(predicted == actual),
    mae = mean(abs(predicted - actual)),
    wkappa = compute_weighted_kappa(actual, predicted),
    .groups = "drop")

rq1_item_summary <- rq1_item |>
  group_by(survey, survey_label, method, method_label,
           method_type, tier) |>
  summarise(
    accuracy = round(mean(accuracy), 3),
    mae = round(mean(mae), 3),
    wkappa = round(mean(wkappa), 3),
    .groups = "drop")
```

```{r}
#| label: tbl-rq1
#| tbl-cap: "RQ1: Item-Level Metrics — All Methods × Tiers (5-Fold CV). Wκ = quadratic weighted kappa."

rq1_item_summary |>
  select(Survey = survey_label, Type = method_type,
         Method = method_label, Tier = tier,
         Acc = accuracy, MAE = mae, `Wκ` = wkappa) |>
  arrange(Survey, Type, Method, Tier) |>
  knitr::kable(align = "lllcccc")
```

## Figure: Tier Effect on Item Recovery

```{r}
#| label: fig-rq1-tiers
#| fig-cap: "RQ1: Item-Level Performance by Information Tier. All methods, both surveys. Upward trajectory from S to P2 confirms psychometric context improves item recovery."
#| fig-width: 10
#| fig-height: 5

ggplot(rq1_item_summary,
       aes(x = tier, y = wkappa,
           color = method_label,
           group = method_label,
           linetype = method_type)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ survey_label, scales = "free_y") +
  scale_color_viridis_d(option = "D", end = 0.85) +
  labs(x = "Information Tier",
       y = "Weighted Kappa",
       color = "Method", linetype = "Type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

\newpage

# RQ2: Construct-Level Recovery

## Person-Level θ Matching

```{r}
#| label: rq2-matching

# ── Match imputed θ to gold θ for test respondents ───────────
rq2_person <- pmap_dfr(
  imputed_fits |>
    filter(!map_lgl(scores, is.null)) |>
    select(survey, fold_id, method, method_label,
           method_type, tier, scores),
  function(survey, fold_id, method, method_label,
           method_type, tier, scores) {
    gold_scores <- gold_fits |>
      filter(survey == !!survey,
             fold_id == !!fold_id) |>
      pull(scores) |> pluck(1)
    if (is.null(gold_scores)) return(NULL)
    test_ids <- folds_list[[survey]] |>
      filter(fold_id == !!fold_id, split == "test") |>
      pull(row_id)
    gold_scores |>
      filter(row_id %in% test_ids) |>
      rename(theta_gold = theta, fs_gold = fs) |>
      inner_join(
        scores |> rename(theta_imp = theta, fs_imp = fs),
        by = "row_id") |>
      mutate(survey = survey, fold_id = fold_id,
             method = method, method_label = method_label,
             method_type = method_type, tier = tier)
  }
) |>
  mutate(survey_label = if_else(survey == "lapop",
    "LAPOP Mexico", "Latinobarómetro Colombia"))

cat("RQ2 person-level:", nrow(rq2_person), "rows\n")
```

## Correlation and Slope

```{r}
#| label: rq2-metrics

rq2_metrics <- rq2_person |>
  group_by(survey, survey_label, method, method_label,
           method_type, tier, fold_id) |>
  summarise(
    r = cor(theta_gold, theta_imp, use = "complete.obs"),
    slope = coef(lm(theta_imp ~ theta_gold))[["theta_gold"]],
    .groups = "drop")

rq2_summary <- rq2_metrics |>
  group_by(survey_label, method_type, method_label, tier) |>
  summarise(
    r = round(mean(r, na.rm = TRUE), 4),
    slope = round(mean(slope, na.rm = TRUE), 4),
    .groups = "drop")
```

```{r}
#| label: tbl-rq2
#| tbl-cap: "RQ2: Construct Recovery — Correlation (r) and Regression Slope of θ_imputed on θ_gold. r = rank preservation; slope = scale fidelity (1.0 = perfect, <1 = compression)."

rq2_summary |>
  select(Survey = survey_label, Type = method_type,
         Method = method_label, Tier = tier,
         r, Slope = slope) |>
  arrange(Survey, Tier, desc(r)) |>
  knitr::kable(align = "lllccc")
```

## Figure: θ Recovery Scatterplots

```{r}
#| label: fig-rq2-scatter
#| fig-cap: "RQ2: Construct Recovery — θ_gold vs. θ_imputed for Claude P2 and mixgb S (Fold 1). Solid line = regression fit; dashed = identity. Deviation from slope = 1 indicates compression or expansion."
#| fig-width: 10
#| fig-height: 5

scatter_data <- rq2_person |>
  filter(fold_id == 1,
         (method == "claude_sonnet" & tier == "P2") |
         (method == "mixgb" & tier == "S")) |>
  mutate(label = paste(method_label,
                        if_else(method_type == "LLM",
                                "(P2)", "(S)")))

ggplot(scatter_data,
       aes(x = theta_gold, y = theta_imp)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50",
              linewidth = 0.5) +
  geom_point(alpha = 0.15, size = 0.8,
             color = viridis(3)[2]) +
  geom_smooth(method = "lm", se = FALSE,
              color = viridis(3)[1],
              linewidth = 0.9) +
  facet_grid(label ~ survey_label, scales = "free") +
  labs(x = expression(theta[gold]),
       y = expression(theta[imputed])) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

## Figure: Correlation by Method and Tier

```{r}
#| label: fig-rq2-r
#| fig-cap: "RQ2: Rank Preservation (r) by Method and Tier. All methods on both surveys."
#| fig-width: 10
#| fig-height: 5

ggplot(rq2_summary,
       aes(x = tier, y = r,
           color = method_label,
           group = method_label,
           linetype = method_type)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ survey_label, scales = "free_y") +
  scale_color_viridis_d(option = "D", end = 0.85) +
  labs(x = "Information Tier",
       y = "Correlation (r) with Gold Standard θ",
       color = "Method", linetype = "Type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```

\newpage


# RQ3: Item Measurement Error

RQ3 asks: what does imputation do to the measurement model's
item parameters? We refit IRT-GRM on four conditions per fold
(gold, deletion, Claude P2, mixgb S), extract discrimination
(a) and raw intercept (d) parameters, and decompose per-item
MSE into Bias² + Variance following Biemer (2010). Item-level
response variance is also compared across conditions.

## Extract IRT Item Parameters

```{r}
#| label: rq3-extract

tic("IRT parameter extraction")
rq3_irt_params <- map_dfr(c("lapop", "lb"), function(s) {
  cfg <- survey_config[[s]]
  map_dfr(1:5, function(fi) {
    cat("  Extracting:", s, "fold", fi, "\n")

    # Gold: full fold (training + test, no masking)
    gold_df <- folds_list[[s]] |>
      filter(fold_id == fi) |>
      select(row_id, all_of(cfg$trust_items)) |>
      mutate(across(all_of(cfg$trust_items), as.numeric))
    gold_p <- fit_irt_extract(gold_df, cfg) |>
      mutate(condition = "Gold")

    # Deletion: training only (listwise deletion of test fold)
    del_df <- folds_list[[s]] |>
      filter(fold_id == fi, split == "train") |>
      select(row_id, all_of(cfg$trust_items)) |>
      mutate(across(all_of(cfg$trust_items), as.numeric))
    del_p <- fit_irt_extract(del_df, cfg) |>
      mutate(condition = "Deletion")

    # Claude Sonnet P2 (best LLM configuration)
    claude_df <- reconstruct_full_data(
      s, fi, "claude_sonnet", "P2")
    claude_p <- fit_irt_extract(claude_df, cfg) |>
      mutate(condition = "Claude P2")

    # mixgb S (traditional benchmark — fair comparison tier)
    mixgb_df <- reconstruct_full_data(s, fi, "mixgb", "S")
    mixgb_p <- if (!is.null(mixgb_df)) {
      fit_irt_extract(mixgb_df, cfg) |>
        mutate(condition = "mixgb S")
    } else {
      tibble(item = character(), param = character(),
             value = numeric(), condition = character())
    }

    bind_rows(gold_p, del_p, claude_p, mixgb_p) |>
      mutate(survey = s, fold_id = fi)
  })
}) |>
  mutate(
    survey_label = if_else(survey == "lapop",
      "LAPOP Mexico", "Latinobarómetro Colombia"),
    condition = factor(condition,
      levels = c("Gold", "Deletion", "Claude P2", "mixgb S"))
  )
toc()

cat("RQ3 IRT params:", nrow(rq3_irt_params), "rows\n")
```

## Discrimination (a): Biemer Decomposition

```{r}
#| label: rq3-discrim

# ── Per item × fold: Δa for all conditions vs Gold ───────────
rq3_delta_a <- rq3_irt_params |>
  filter(param == "a") |>
  select(survey_label, item, fold_id, condition, value) |>
  pivot_wider(names_from = condition,
              values_from = value) |>
  mutate(
    delta_claude = `Claude P2` - Gold,
    delta_del    = Deletion - Gold,
    delta_mixgb  = `mixgb S` - Gold
  )

# ── Helper: Biemer decomposition for any delta column ────────
# Returns per-item Bias, Bias², Variance, MSE, and % Syst
biemer_decompose <- function(data, delta_col, config_label) {
  data |>
    group_by(survey_label, item) |>
    summarise(
      gold_a   = round(mean(Gold), 3),
      bias     = round(mean(.data[[delta_col]]), 4),
      bias_sq  = mean(.data[[delta_col]])^2,
      variance = var(.data[[delta_col]]),
      mse      = mean(.data[[delta_col]]^2),
      pct_syst = round(
        if_else(mean(.data[[delta_col]]^2) > 0,
                mean(.data[[delta_col]])^2 /
                  mean(.data[[delta_col]]^2) * 100,
                0), 1),
      .groups = "drop"
    ) |>
    mutate(config = config_label) |>
    arrange(survey_label, desc(mse))
}

# ── Apply to all three conditions ─────────────────────────────
rq3_biemer_claude <- biemer_decompose(
  rq3_delta_a, "delta_claude", "Claude P2")
rq3_biemer_del <- biemer_decompose(
  rq3_delta_a, "delta_del", "Deletion")
rq3_biemer_mixgb <- biemer_decompose(
  rq3_delta_a, "delta_mixgb", "mixgb S")
```

## Table: Item-Level Biemer — Claude P2 vs mixgb S

```{r}
#| label: tbl-rq3-discrim
#| tbl-cap: "RQ3: IRT Discrimination (a) — Biemer Decomposition. Claude P2 and mixgb S vs Gold Standard. Bias = mean Δa across 5 folds. MSE = Bias² + Variance."

# ── Side-by-side item-level table ─────────────────────────────
rq3_item_compare <- rq3_biemer_claude |>
  select(survey_label, item, gold_a,
         bias_c = bias, mse_c = mse, pct_c = pct_syst) |>
  left_join(
    rq3_biemer_mixgb |>
      select(survey_label, item,
             bias_m = bias, mse_m = mse, pct_m = pct_syst),
    by = c("survey_label", "item")
  ) |>
  arrange(survey_label, desc(gold_a))

rq3_item_compare |>
  mutate(
    bias_c = sprintf("%+.4f", bias_c),
    mse_c  = sprintf("%.4f", mse_c),
    bias_m = sprintf("%+.4f", bias_m),
    mse_m  = sprintf("%.4f", mse_m)
  ) |>
  select(Survey = survey_label, Item = item,
         `Gold a` = gold_a,
         `Claude Bias` = bias_c, `Claude MSE` = mse_c,
         `Claude %Sys` = pct_c,
         `mixgb Bias` = bias_m, `mixgb MSE` = mse_m,
         `mixgb %Sys` = pct_m) |>
  knitr::kable(align = "llccccccc")
```

## Table: Aggregate MSE Summary (Three Conditions)

```{r}
#| label: tbl-rq3-discrim-agg
#| tbl-cap: "RQ3: Discrimination MSE Summary — Claude P2, mixgb S, and Deletion. Mean across items per survey."

bind_rows(
  rq3_biemer_claude, rq3_biemer_mixgb, rq3_biemer_del
) |>
  group_by(survey_label, config) |>
  summarise(
    mean_bias    = round(mean(bias), 4),
    mean_bias_sq = round(mean(bias_sq), 6),
    mean_var     = round(mean(variance), 6),
    mean_mse     = round(mean(mse), 6),
    mean_pct     = round(mean(pct_syst), 1),
    .groups = "drop"
  ) |>
  mutate(
    config = factor(config,
      levels = c("Claude P2", "mixgb S", "Deletion"))
  ) |>
  select(Survey = survey_label, Condition = config,
         Bias = mean_bias, `Bias²` = mean_bias_sq,
         Variance = mean_var, MSE = mean_mse,
         `% Syst` = mean_pct) |>
  arrange(Survey, Condition) |>
  knitr::kable(align = "llccccc")
```

## Figure: Discrimination Shift Dumbbell (Claude P2 vs mixgb S)

```{r}
#| label: fig-rq3-dumbbell
#| fig-cap: "RQ3: IRT Discrimination Shift by Item — Claude P2 vs mixgb S. Each segment connects the two methods' mean Δa (condition − Gold) for the same item. Vertical dashed line = gold standard (no distortion)."
#| fig-width: 10
#| fig-height: 6

# ── Build dumbbell data ───────────────────────────────────────
dumbbell_df <- rq3_biemer_claude |>
  select(survey_label, item, gold_a,
         bias_claude = bias) |>
  left_join(
    rq3_biemer_mixgb |>
      select(survey_label, item, bias_mixgb = bias),
    by = c("survey_label", "item")
  )

# ── Viridis colors for the two methods ────────────────────────
method_colors <- c("Claude P2" = viridis(4)[3],
                   "mixgb S"   = viridis(4)[1])

ggplot(dumbbell_df, aes(y = reorder(item, gold_a))) +
  # Zero reference line (gold standard = no distortion)
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50") +
  # Connecting segment between methods per item
  geom_segment(
    aes(x = bias_claude, xend = bias_mixgb,
        yend = reorder(item, gold_a)),
    color = "grey60", linewidth = 0.6
  ) +
  # Claude P2 points
  geom_point(aes(x = bias_claude, color = "Claude P2"),
             size = 3) +
  # mixgb S points
  geom_point(aes(x = bias_mixgb, color = "mixgb S"),
             size = 3) +
  # Labels: Claude P2
  geom_text(
    aes(x = bias_claude,
        label = sprintf("%+.3f", bias_claude)),
    hjust = -0.3, size = 2.5, color = viridis(4)[3]
  ) +
  # Labels: mixgb S
  geom_text(
    aes(x = bias_mixgb,
        label = sprintf("%+.3f", bias_mixgb)),
    hjust = 1.3, size = 2.5, color = viridis(4)[1]
  ) +
  facet_wrap(~ survey_label, scales = "free") +
  scale_color_manual(values = method_colors, name = NULL) +
  labs(
    x = expression(Delta * "a (Condition − Gold)"),
    y = NULL,
    title = "Discrimination shift: Claude P2 vs mixgb S"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Figure: Biemer Decomposition — Both Methods

```{r}
#| label: fig-rq3-biemer
#| fig-cap: "RQ3: Biemer Decomposition of Discrimination MSE — Claude P2 vs mixgb S. Dark fill = Bias² (systematic), light fill = Variance (random). Labels show total MSE and % systematic."
#| fig-width: 10
#| fig-height: 7

# ── Combine both imputation methods for stacked bar ───────────
biemer_both <- bind_rows(rq3_biemer_claude, rq3_biemer_mixgb)

biemer_long <- biemer_both |>
  pivot_longer(cols = c(bias_sq, variance),
               names_to = "component",
               values_to = "comp_value") |>
  mutate(
    component = factor(component,
      levels = c("bias_sq", "variance"),
      labels = c("Bias²", "Variance")),
    config = factor(config,
      levels = c("Claude P2", "mixgb S"))
  )

ggplot(biemer_long,
       aes(x = comp_value,
           y = reorder(item, comp_value),
           fill = component)) +
  geom_col(width = 0.6, position = "stack") +
  geom_text(
    data = biemer_both,
    aes(x = mse,
        y = reorder(item, mse),
        label = paste0(sprintf("%.4f", mse),
                       "  (", pct_syst, "%)"),
        fill = NULL),
    hjust = -0.05, size = 2.5
  ) +
  facet_grid(config ~ survey_label, scales = "free") +
  scale_fill_manual(
    values = c("Bias²"    = viridis(3)[2],
               "Variance" = "grey75"),
    name = NULL
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.40))) +
  labs(
    x = "MSE of Discrimination (a)",
    y = NULL,
    title = "Biemer decomposition: MSE = Bias² + Variance"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Item-Level Response Variance

```{r}
#| label: rq3-item-var

# ── Per-item variance: Gold vs Deletion vs Claude P2
#    vs mixgb S ────────────────────────────────────────────────
rq3_item_var <- map_dfr(c("lapop", "lb"), function(s) {
  cfg <- survey_config[[s]]
  items <- cfg$trust_items
  map_dfr(1:5, function(fi) {
    gold_data <- folds_list[[s]] |>
      filter(fold_id == fi) |>
      select(all_of(items)) |>
      mutate(across(everything(), as.numeric))
    del_data <- folds_list[[s]] |>
      filter(fold_id == fi, split == "train") |>
      select(all_of(items)) |>
      mutate(across(everything(), as.numeric))
    claude_data <- reconstruct_full_data(
      s, fi, "claude_sonnet", "P2") |>
      select(all_of(items))
    mixgb_recon <- reconstruct_full_data(
      s, fi, "mixgb", "S")
    mixgb_data <- if (!is.null(mixgb_recon)) {
      mixgb_recon |> select(all_of(items))
    } else { NULL }
    map_dfr(items, function(it) {
      tibble(
        survey = s, fold_id = fi, item = it,
        var_gold   = var(gold_data[[it]], na.rm = TRUE),
        var_del    = var(del_data[[it]], na.rm = TRUE),
        var_claude = var(claude_data[[it]], na.rm = TRUE),
        var_mixgb  = if (!is.null(mixgb_data)) {
          var(mixgb_data[[it]], na.rm = TRUE)
        } else { NA_real_ }
      )
    })
  })
}) |>
  mutate(
    survey_label = if_else(survey == "lapop",
      "LAPOP Mexico", "Latinobarómetro Colombia"),
    pct_claude = round((var_claude - var_gold) /
                        var_gold * 100, 2),
    pct_del    = round((var_del - var_gold) /
                        var_gold * 100, 2),
    pct_mixgb  = round((var_mixgb - var_gold) /
                        var_gold * 100, 2)
  )

# ── Summary across folds ─────────────────────────────────────
rq3_var_summary <- rq3_item_var |>
  group_by(survey_label, item) |>
  summarise(
    var_gold   = round(mean(var_gold), 3),
    var_claude = round(mean(var_claude), 3),
    var_del    = round(mean(var_del), 3),
    var_mixgb  = round(mean(var_mixgb, na.rm = TRUE), 3),
    pct_claude = round(mean(pct_claude), 2),
    pct_del    = round(mean(pct_del), 2),
    pct_mixgb  = round(mean(pct_mixgb, na.rm = TRUE), 2),
    .groups = "drop"
  )
```

```{r}
#| label: tbl-rq3-var
#| tbl-cap: "RQ3: Item-Level Response Variance — Gold vs Deletion vs Claude P2 vs mixgb S. pct = % change from gold. Negative = variance shrinkage."

rq3_var_summary |>
  arrange(survey_label, item) |>
  select(Survey = survey_label, Item = item,
         `Gold` = var_gold,
         `Claude` = var_claude, `%C` = pct_claude,
         `mixgb` = var_mixgb, `%M` = pct_mixgb,
         `Del` = var_del, `%D` = pct_del) |>
  knitr::kable(align = "llccccccc")
```

## Figure: Item Variance Shrinkage

```{r}
#| label: fig-rq3-var
#| fig-cap: "RQ3: Item-Level Variance Change vs Gold. Claude P2 and mixgb S both shrink item response variance; the method with less shrinkage preserves the measurement model's error structure more faithfully."
#| fig-width: 10
#| fig-height: 6

# ── Long format for plotting both methods ─────────────────────
var_long <- rq3_var_summary |>
  pivot_longer(
    cols = c(pct_claude, pct_mixgb),
    names_to = "method",
    values_to = "pct_change"
  ) |>
  mutate(
    method = recode(method,
      pct_claude = "Claude P2",
      pct_mixgb  = "mixgb S"),
    method = factor(method,
      levels = c("Claude P2", "mixgb S"))
  )

ggplot(var_long,
       aes(x = pct_change,
           y = reorder(item, pct_change),
           color = method)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey50") +
  geom_point(size = 2.5,
             position = position_dodge(0.4)) +
  facet_wrap(~ survey_label, scales = "free") +
  scale_color_manual(
    values = c("Claude P2" = viridis(4)[3],
               "mixgb S"   = viridis(4)[1]),
    name = NULL
  ) +
  labs(
    x = "% change in item variance (vs Gold)",
    y = NULL,
    title = "Item-level variance shrinkage: Claude P2 vs mixgb S"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

\newpage

# Save All Results

```{r}
#| label: save

saveRDS(rq1_item,
  file.path(results_dir, "rq1_item_metrics.rds"))
saveRDS(rq2_person,
  file.path(results_dir, "rq2_person_recovery.rds"))

save(
  # Model fits
  gold_fits, deletion_fits, imputed_fits,
  # RQ1
  rq1_item, rq1_item_summary,
  # RQ2
  rq2_person, rq2_metrics, rq2_summary,
  # RQ3
  rq3_irt_params, rq3_delta_a,
  rq3_biemer_claude, rq3_biemer_mixgb, rq3_biemer_del,
  rq3_item_var, rq3_var_summary,
  file = file.path(results_dir, "evaluation_rq.RData"))

cat("Saved: evaluation_rq.RData,",
    "rq1_item_metrics.rds,",
    "rq2_person_recovery.rds\n")

tibble(
  Object = c("gold_fits", "deletion_fits", "imputed_fits",
             "rq1_item", "rq2_person", "rq2_summary",
             "rq3_irt_params", "rq3_biemer_claude",
             "rq3_biemer_mixgb", "rq3_item_var"),
  Rows = c(nrow(gold_fits), nrow(deletion_fits),
           nrow(imputed_fits), nrow(rq1_item),
           nrow(rq2_person), nrow(rq2_summary),
           nrow(rq3_irt_params), nrow(rq3_biemer_claude),
           nrow(rq3_biemer_mixgb), nrow(rq3_item_var))
) |> knitr::kable()
```
