---
title: "L2L Framework: Prompt Context Engineering"
subtitle: "Design Rationale, Tier Architecture, and Production Decision Log"
author: "Kevin Linares"
date: today
format:
  pdf:
    colorlinks: true
    fig-pos: "H"
    geometry: margin=1in
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  echo: true
  warning: false
  message: false
  cache: false
editor:
  markdown:
    wrap: 72
---

# Overview

This document describes the prompt construction process for the
Latent-to-Language (L2L) item imputation framework. It records the
design rationale, tier structure, prompt architecture, iterative
development decisions, and production results for peer review. No API
calls are made here --- all prompts are printed from
`item_prompt_functions.R` for inspection.

The L2L framework tests whether translating psychometric model
*conclusions* --- not raw parameters --- into natural language prompts
enables LLMs to impute masked survey item responses competitively
with tree-based methods. The production design uses a
**model-then-adjust** paradigm: IRT/CFA model predictions serve as
starting points that LLMs refine based on residual patterns,
correlated-item signals, and demographic probability context.

**Key design principle:** Prompts communicate model conclusions in
natural language. The LLM is a reasoner, not a calculator. It never
sees discrimination parameters, factor loadings, Q3 matrices, or
threshold tables.

All prompt construction functions are in `item_prompt_functions.R`,
a self-contained R script sourced by Script 04:

```r
source("D:/repos/UMD_classes_code/TSE_II_SURV721/code/item_prompt_functions.R")
```

# Tier Structure

## Output Format

Batch prompting with model-specific batch sizes: Claude Sonnet = 50,
Gemma 27B = 25, Maverick = 25. Each respondent block shows
demographics, observed items with verbal labels and `[MISSING]`
markers, and tier-specific psychometric context. Output is one line
per respondent: `row_id: item1=value, item2=value`.

## LLM Tiers

- **S (Demographics only):** Demographics + observed trust items as
  verbal labels + `[MISSING]` markers. Pure world-knowledge
  prediction. No model predictions, no psychometric context.

- **P1 (CFA augmented):** S + per-respondent CFA factor score
  (percentile and verbal label) + per-masked-item CFA prediction
  anchor + 2 most correlated observed items with CFA residuals +
  overall residual pattern summary.

- **P2 (IRT augmented):** S + per-masked-item IRT modal prediction
  with verbal confidence level and top-2 category probabilities +
  2 most correlated observed items with IRT residuals + overall IRT
  residual pattern + demographic probability context
  (P(high trust | group) vs. population).

## Traditional Method Tiers (Script 03)

- **S:** Demographics + observed trust items (masked as NA)
- **P1:** S + CFA factor score (from partial data)
- **P2:** S + IRT $\theta$ (from partial data, no SE)

## Demographics (Screened by Script 02)

Demographics were screened using eta-squared with both IRT theta and
CFA factor score (threshold: $\eta^2 \geq 0.005$). Each survey
retains only demographics that pass screening:

- **LAPOP Mexico:** age\_cat, Edu, Urban, Employment (male dropped,
  $\eta^2 = 0.0005$ with $\theta$, $0.0007$ with CFA)
- **LB Colombia:** age\_cat, Edu, Employment (male and Urban dropped)

# Psychometrician Persona

The system prompt uses a specific persona: a survey psychometrician
with a PhD from JPSM at the University of Maryland, specializing in
survey data science. The persona establishes domain expertise in
institutional trust measurement for Latin American populations.

The system prompt frames the task as **refinement**, not binary
keep/adjust:

> "Your job is to REFINE these predictions using information the
> model cannot capture. The model only sees numerical patterns ---
> you can see response consistency, demographic context, and
> institutional meaning."

## Reasoning Rules (System Prompt)

1. **Residual pattern is the strongest signal.** If the respondent
   consistently rates above or below the model on observed items,
   adjust in that direction.

2. **Correlated items provide direct evidence.** If the two most
   related observed items are both above (or below) the model,
   adjust in the same direction.

3. **Demographic context is a sanity check.** Respondent-level
   patterns always outweigh group averages.

4. **Domain knowledge matters.** Trust in armed forces is
   substantively different from trust in the judiciary.

5. **Low confidence = your reasoning matters most.** When the model's
   top category has $< 30\%$ probability, do not simply repeat the
   prediction.

6. **Halo effects.** If a respondent rates all observed institutions
   consistently high or low, missing items almost certainly follow.

7. **Integer output.** Predictions must be on the original scale.

# The Model-Then-Adjust Paradigm

## Evolution from Raw Parameters

Early designs (V1) provided raw psychometric parameters: CFA
standardized loadings, IRT discrimination and threshold tables in
mirt format, Q3 residual correlation matrices, DIF flags from
lordif, measurement invariance results, and theta regression
coefficients. LLMs received no model predictions --- they were asked
to compute predictions from these parameters.

**Result:** Consistently below IRT baseline. LLMs cannot mentally
evaluate the GRM probability function or convert discrimination
parameters into item-level predictions.

## Current Design (V2)

The model-then-adjust paradigm provides the statistical model's
conclusion as a starting point. For each masked item, the LLM
receives:

1. **Anchor prediction:** The IRT modal category (P2) or CFA
   rounded prediction (P1).

2. **Confidence signal (P2 only):** Verbal label (HIGH/MODERATE/LOW)
   with top-2 category probabilities. Directive framing:
   - HIGH: "model is fairly certain"
   - MODERATE: "use residual pattern to refine"
   - LOW: "model is uncertain, your reasoning matters most"

3. **Correlated items (2 per masked item):** The two observed items
   most correlated with the masked item (from CFA model-implied
   correlation matrix), showing observed value, model expected value,
   and residual with explicit directional implication:
   - "+0.8 above $\to$ suggests this item should also be ABOVE model"
   - "-0.5 below $\to$ suggests this item should also be BELOW model"

4. **Overall residual pattern:** One-sentence prescriptive summary:
   - "This respondent consistently rates ABOVE the model (+0.5 avg)
     $\to$ adjust predictions UPWARD"

5. **Demographic probability context (P2 only):** Per-item
   model-based probability: "For your group, P(high trust on
   armed\_forces) = 48\% vs. population 43\%."

## What Was Dropped from V1

- IRT parameter tables (discrimination $a$, thresholds $b$)
- CFA standardized loadings and SE
- Q3 local dependency matrix
- Conditional expectation $E[X_M | X_O]$
- DIF flags from lordif
- Measurement invariance non-invariant items
- Theta-demographic regression coefficients and $R^2$
- Theta with SE and percentile as standalone prompt line
- Chain-of-thought reasoning tags

All dropped elements either added noise (LLMs cannot compute from
raw parameters) or were replaced by more interpretable alternatives
(DIF flags $\to$ demographic probability context; theta regression
$\to$ group-level P(high trust)).

# Few-Shot Examples

Three training-fold exemplars per tier (P1, P2), spanning the latent
trust spectrum. No few-shot for S tier (pure world knowledge).

1. **Low trust, low residual $\to$ KEEP.** Model confidence is high
   and the respondent's pattern is consistent with predictions.

2. **Moderate trust, strong positive residual $\to$ ADJUST UP.** The
   respondent consistently rates above the model, and uncertainty is
   high.

3. **High trust, negative residual $\to$ ADJUST DOWN.** Despite high
   overall trust, this item's pattern suggests lower.

Each example shows the complete prompt block for one respondent
(demographics, observed items, residual summary, model prediction
with confidence, correlated items) and the correct output line.
Exemplars are selected from the training fold: low from the bottom
tertile of latent scores, moderate from the middle, high from the
top.

**Reduction from 5 to 3 examples:** The original 5-exemplar design
included "moderate residual --- small adjustment" and "extreme
latent score --- edge case." Testing showed these two added tokens
without improving accuracy. The three remaining examples cover the
decision space: keep, adjust up, adjust down.

# Multiple Imputation Design (m = 5)

## Rationale

LLM predictions at P2 are treated as multiple imputations, mirroring
the $m = 5$ majority-vote design used for mixgb and miceRanger in
Script 03. Each fold $\times$ tier $\times$ model combination is
run 5 times at graduated temperatures to introduce controlled
stochasticity:

| Run | Temperature |
|-----|-------------|
| 1   | 0.01        |
| 2   | 0.10        |
| 3   | 0.20        |
| 4   | 0.30        |
| 5   | 0.40        |

## Benefits

1. **Parse failure recovery:** If a respondent fails to parse in 1
   of 5 runs, the other 4 provide predictions.

2. **Prediction stabilization:** Majority vote across 5 runs reduces
   stochastic variance.

3. **Methodological symmetry:** Both traditional and LLM methods use
   $m = 5$ with majority vote.

4. **Between-imputation variance:** Raw per-run results are saved for
   MI variance analysis (RQ1 latent variable recovery).

## Cost

Approximately 5$\times$ the single-run cost. Claude Sonnet at batch
50: $\sim$\$37 for all 5 folds $\times$ 3 tiers $\times$ 5 runs.
Gemma and Maverick through OpenRouter are substantially cheaper.

# Measurement Models

## CFA (for P1)

Per fold, the CFA provides factor scores from partial data via
`lavPredict()` with FIML for missing items. Factor scores are
expressed as percentiles and verbal labels (VERY LOW through VERY
HIGH) relative to the training distribution.

CFA predictions for masked items use the linear model:
$\hat{X}_j = \hat{\tau}_j + \hat{\lambda}_j \hat{\eta}$, clamped
to the scale range and rounded.

CFA residuals (observed $-$ expected) for observed items provide
the respondent-level consistency signal.

## GRM/IRT (for P2)

The Graded Response Model provides $\theta$ from partial data via
`fscores()` (EAP method). For each masked item, `probtrace()`
computes the full category probability vector, from which the modal
category, confidence level, and top-2 probabilities are derived.

IRT residuals (observed $-$ $E[X_j | \hat{\theta}]$) for observed
items are the primary adjustment signal.

For LB (descending scale), IRT reverse-coding ensures higher
$\theta$ = higher trust. Probabilities are reversed back to the
original scale in prompts.

## Demographic Profiling (Script 02)

Script 02 computes model-based probabilities per demographic group:
$P(X_j \geq k | \bar{\theta}_{\text{group}})$ using the IRT model.
These replace the DIF flags and theta regression from V1 with
directly actionable probability comparisons. In the prompt:
"For your group, P(high trust on armed\_forces) = 48\% vs. population
43\%."

## Baselines

- **IRT baseline:** $\arg\max_k P(X_j = k | \hat{\theta})$ from
  partial-data theta
- **CFA baseline:** $\text{round}(\hat{\tau}_j + \hat{\lambda}_j
  \hat{\eta})$ clamped to scale

Both are computed from partial data and recorded alongside every
prediction in the results tibble.

# Setup

```{r}
#| label: setup

library(lavaan)
library(mirt)
library(tidyverse)

set.seed(721)

code_dir    <- "D:/repos/UMD_classes_code/TSE_II_SURV721/code"
results_dir <- "D:/repos/UMD_classes_code/TSE_II_SURV721/results"

# Source item imputation prompt functions
source(file.path(code_dir, "item_prompt_functions.R"))

# Load fold data and psychometric objects
load(file.path(results_dir, "fold_data.RData"))
load(file.path(results_dir, "cfa_reporting.RData"))
load(file.path(results_dir, "irt_models.RData"))

cat("Loaded: folds_list, cfa_list, irt_list, masking_tbl,\n")
cat("  cfa_params_tbl, irt_params_tbl, q3_pairs_tbl,\n")
cat("  demo_profile_tbl, irt_models\n")

# Helper: wrap long lines for PDF output
wrap_prompt <- function(text, width = 78) {
  lines <- str_split(text, "\n")[[1]]
  map_chr(lines, function(line) {
    if (nchar(line) <= width) return(line)
    str_wrap(line, width = width, exdent = 2)
  }) |> str_c(collapse = "\n")
}
```

# Survey Configuration

```{r}
#| label: config

config <- list(
  lapop = list(
    survey_label    = "a nationally representative survey conducted in Mexico in 2023",
    demo_vars       = c("age_cat", "Edu", "Urban", "Employment"),
    trust_items     = c(
      "justice_system", "electoral_tribunal", "armed_forces",
      "legislature", "public_ministry", "police", "auditor",
      "political_parties", "supreme_court", "municipality",
      "media", "elections"
    ),
    factor_names    = "Trust",
    trust_ascending = TRUE,
    item_scale_desc = "1 = no trust at all, 7 = a lot of trust",
    cfa_syntax = '
      Trust =~ justice_system + electoral_tribunal + armed_forces +
               legislature + public_ministry + police + auditor +
               political_parties + supreme_court + municipality +
               media + elections
      armed_forces ~~ legislature
      public_ministry ~~ auditor
    '
  ),
  lb = list(
    survey_label    = "a nationally representative survey conducted in Colombia in 2023",
    demo_vars       = c("age_cat", "Edu", "Employment"),
    trust_items     = c(
      "Congress", "Natl_Govt", "Judiciary", "Parties",
      "Electoral_Inst", "President"
    ),
    factor_names    = "Political",
    trust_ascending = FALSE,
    item_scale_desc = "1 = a lot of trust, 4 = no trust at all",
    cfa_syntax = '
      Political =~ Congress + Natl_Govt + Judiciary + Parties +
                   Electoral_Inst + President
      Natl_Govt ~~ President
    '
  )
)

cat("LAPOP:", length(config$lapop$trust_items), "items,",
    config$lapop$item_scale_desc, "\n")
cat("LB:   ", length(config$lb$trust_items), "items,",
    config$lb$item_scale_desc, "\n")
```

# System Prompt

The system prompt is tier-invariant — the same persona and reasoning
rules apply across all tiers. Only the user prompt content changes.

```{r}
#| label: system-prompt

system_prompt <- build_system_prompt(config$lapop)
cat(wrap_prompt(system_prompt))
```

\newpage

# Prompt Examples

## Tier S (No Model)

```{r}
#| label: example-tier-s

# Pick one test respondent from LAPOP fold 1
test_row <- folds_list$lapop |>
  filter(fold_id == 1, split == "test") |>
  slice(1)

masked <- c("armed_forces", "legislature")

info_s <- list(
  rid          = test_row$row_id,
  masked_items = masked
)

s_block <- build_respondent_block_s(info_s, test_row, config$lapop)
cat(wrap_prompt(s_block))
```

\newpage

# Key Design Decisions

## Model-Then-Adjust vs. Predict-From-Scratch

LLMs receiving the IRT prediction as an anchor consistently
outperformed LLMs receiving only model parameters. LLMs cannot
mentally evaluate the GRM probability function, but they can reason
about whether to adjust a given prediction based on contextual
signals. This is the core insight of the L2L framework:
**translate model conclusions, not model parameters.**

## Conclusions Over Parameters

The V1 design included IRT discrimination tables ($a_j$, $b_{jk}$),
CFA loadings ($\lambda_j$), Q3 matrices, DIF flags, and regression
coefficients. LLMs treated these as noise. The V2 design replaced
all parameters with natural language conclusions:

| V1 (Parameters)             | V2 (Conclusions)                     |
|-----------------------------|--------------------------------------|
| $a_j = 3.30$, $b_{jk}$ table | "Model prediction: 5, Confidence: HIGH (5 at 42%, 4 at 28%)" |
| Q3 matrix                   | "Most informative: police = 6 (+0.8 above $\to$ adjust up)" |
| DIF flag: "media shows DIF by Edu" | "For your group, P(high trust on media) = 35% vs. population 43%" |
| $\theta \sim \text{Edu}$: $b = -0.15$ | (Absorbed into demographic probability context) |
| Residual: $-0.32$           | "Respondent rates BELOW model $\to$ adjust DOWNWARD" |

## Prescriptive vs. Descriptive Prompts

V1 prompts were descriptive: "This respondent is +0.5 above model
expectations." V2 prompts are prescriptive: "This respondent
consistently rates ABOVE the model (+0.5 avg) $\to$ adjust
predictions UPWARD." Testing showed that prescriptive framing does
not significantly improve accuracy over descriptive framing for
Claude Sonnet, but it reduces the gap for smaller models (Gemma,
Maverick) by making the adjustment direction explicit.

## No Chain-of-Thought

CoT was tested with both Maverick and Claude Sonnet under both V1
and V2 architectures. In every configuration, CoT degraded accuracy.
Parse rates dropped and predictions were worse. The reasoning step
introduces noise — the LLM talks itself out of correct adjustments.

## No Base Rates

Base rate distributions encourage modal parroting. The IRT model
predictions implicitly encode population-level information. Explicit
base rates are redundant and harmful.

## Three Few-Shot Examples (Reduced from Five)

The five-exemplar design was reduced to three (low/moderate/high
trust) without performance loss. The two dropped exemplars
(moderate residual and extreme latent score) added tokens without
covering distinct adjustment scenarios.

## Batch Size

Testing across sizes (1, 5, 10, 20, 25, 50) showed negligible
accuracy differences for Claude Sonnet (range: 62.3--63.9\% on LB
P2). Claude uses batch 50 for cost efficiency. Gemma and Maverick
use batch 25 due to higher parse failure rates at larger sizes.

## Masking Design

1--4 items per respondent, randomly assigned (`sample(1:4)` count,
`sample(items, size = k)` selection). Produces a continuous
degradation curve. Masking table generated once in Script 02 and
shared across all methods.

## Opus 4.6 vs. Sonnet 4.6

Opus was tested on one fold --- accuracy matched Sonnet at
approximately 15$\times$ the cost. Sonnet is the production model.

# Models Evaluated

## LLMs

- **Claude Sonnet 4.6** (Anthropic API, batch 50): Strongest LLM.
  The only model consistently above IRT baseline at P2 on both
  surveys.
- **Llama 4 Maverick** (Meta, OpenRouter, batch 25): Below IRT at
  P2 but shows the tier ordering (S $<$ P2).
- **Gemma 3 27B** (Google, OpenRouter, batch 25): Weakest LLM.
  Below IRT at P2 on both surveys.

## Traditional Benchmarks

- **mixgb** (XGBoost, $m = 5$ majority vote): Strongest traditional
  benchmark.
- **miceRanger** (Random forest, $m = 5$ majority vote): Below mixgb
  on both surveys.

## Parse Rates (5-Fold Production, m=5 MI)

| Method  | LAPOP S | LAPOP P1 | LAPOP P2 | LB S | LB P1 | LB P2 |
|---------|---------|----------|----------|------|-------|-------|
| Claude  | 90.1%   | 100%     | 100%     | 86.1%| 100%  | 100%  |
| Gemma   | 99.9%   | 96.9%    | 98.1%    | 100% | 96.1% | 98.6% |
| Maverick| 100%    | 100%     | 100%     | 100% | 100%  | 100%  |

Claude's lower S-tier parse rate reflects the absence of the model
anchor — without a structured prediction block, Claude occasionally
produces verbose output. P1 and P2 parse rates are 100% for Claude.

# Production Results (5-Fold CV, m=5 MI)

## Accuracy by Method $\times$ Tier $\times$ Survey

| Method  | LAPOP S | LAPOP P1 | LAPOP P2 | LB S  | LB P1 | LB P2 |
|---------|---------|----------|----------|-------|-------|-------|
| Claude  | .362    | .379     | **.394** | .603  | .603  | **.614** |
| Maverick| .366    | .371     | .367     | .595  | .581  | .592  |
| Gemma   | .375    | .347     | .358     | .589  | .557  | .586  |
| IRT     | ---     | ---      | .388     | ---   | ---   | .600  |
| mixgb   | .382    | .393     | .393     | .566  | .574  | .589  |

Claude P2 exceeds IRT by +0.6pp (LAPOP) and +1.4pp (LB). Claude P2
exceeds mixgb P2 by +0.1pp (LAPOP) and +2.5pp (LB). The advantage
is scale-dependent: larger on the 4-point LB scale.

**Note:** Traditional method results at P1/P2 are reported for
completeness but the fair comparison is LLM P1/P2 vs. traditional S,
since trees at P1/P2 see the complete training response matrix plus
psychometric features — an information advantage that has nothing to
do with the L2L innovation. See Script 05 (H1c) for the fair
comparison.

## Key Findings

1. **Tier ordering (H1a):** Claude shows monotonic S $<$ P1 $<$ P2
   on both surveys. Gemma and Maverick show the ordering on some
   but not all conditions.

2. **Scale dependency:** LLM advantage is larger on the 4-point
   scale (LB) where categorical predictions have higher base rates.
   On the 7-point scale (LAPOP), the IRT baseline is harder to beat.

3. **Model capability threshold:** Only the frontier model (Claude
   Sonnet) consistently exceeds IRT. Gemma 27B and Maverick cannot
   reliably extract the adjustment signal from residual patterns.
   This is itself a finding about the reasoning capability required
   for psychometric context engineering.

4. **Parse rate $\to$ data quality:** MI replication ($m = 5$)
   recovers from parse failures. After majority vote, effective
   coverage is $> 98\%$ for all model $\times$ tier conditions.

# Prompt Engineering Evolution

1. **V1 (Raw parameters):** CFA loadings, IRT discrimination/
   thresholds, Q3, DIF flags, MI results, theta regression, few-shot
   reasoning chains. No model predictions as anchors. Result:
   consistently below IRT baseline.

2. **V2a (Model-then-adjust, passive):** IRT/CFA predictions as
   anchors with "default to keeping unless clear evidence." Five
   few-shot examples. DIF cues and conditional expectations included.
   Result: LLMs anchored too hard --- zero marginal recovery.

3. **V2b (Model-then-adjust, simplified):** Dropped all raw
   parameters, DIF, MI, conditional expectations. Added demographic
   probability context from Script 02. Reduced to 3 few-shot
   examples. Result: Claude matches IRT on LAPOP, exceeds on LB.

4. **V2c (Prescriptive framing):** Changed residual summaries and
   correlated items from descriptive to prescriptive ("$\to$ adjust
   UPWARD"). Changed system prompt from "decide whether to KEEP or
   ADJUST" to "your job is to REFINE." Dropped the $\pm 2$ cap.
   Result: no significant improvement for Claude, marginal
   improvement for smaller models. Adopted as production.

5. **MI replication (m = 5):** Graduated temperature schedule
   (0.01--0.40) across 5 runs with majority vote. Matches traditional
   MI design. Recovers parse failures and stabilizes predictions.

# Limitations

1. **Listwise deletion** reduces samples (LAPOP: 1,430; LB: 1,073).
2. **ML estimation** treats ordinal items as continuous.
3. **Survey weights not used** — evaluation is case-level.
4. **Unidimensional IRT** — Q3 identifies local dependency but does
   not model it.
5. **7-point LAPOP** produces diffuse IRT distributions (max
   $P < 0.25$), limiting the confidence signal.
6. **CFA refit per fold** in Script 04 introduces minor variation
   from Script 02 CFA.
7. **Model-then-adjust** bounds accuracy near IRT — marginal recovery
   is modest (+0.6pp LAPOP, +1.4pp LB for Claude).
8. **Random masking (1--4)** introduces heterogeneity in latent score
   estimation quality.
9. **Cost constraints** — 3 LLMs $\times$ 3 tiers $\times$
   2 surveys $\times$ 5 folds $\times$ 5 MI runs.
10. **Batch size asymmetry** — Claude uses batch 50, others use 25.
    Sensitivity analysis shows negligible effect of batch size on
    accuracy.
11. **S-tier parse rate** — Claude's lower S-tier parse rate (86--90%)
    is a limitation that could bias S-tier results toward parseable
    (potentially easier) cases.

\newpage

# Verify `item_prompt_functions.R`

```{r}
#| label: verify-functions

source_path <- file.path(code_dir, "item_prompt_functions.R")

if (file.exists(source_path)) {
  tryCatch({
    source(source_path, local = TRUE)
    cat("Verified item_prompt_functions.R at:", source_path, "\n")
    cat("File size:", round(file.size(source_path) / 1e3, 1), "KB\n")
    cat("Functions loaded successfully.\n")
  }, error = function(e) {
    cat("ERROR sourcing:", e$message, "\n")
  })
} else {
  cat("WARNING: not found at:", source_path, "\n")
}
```
